---
title: "Multiple Logistic Regression"
author: "SDS 291"
date: "April 13, 2020"
output: slidy_presentation
---

```{r, message=FALSE, warning=FALSE, echo=FALSE}
require(knitr)
opts_chunk$set(eval=TRUE, cache=TRUE)
```

```{r, echo=FALSE, message=FALSE, error=FALSE}
require(Stat2Data)
require(tidyverse)
require(mosaic)
require(mosaicData)
require(lmtest)
data("MedGPA")
```

## Thoughts from HW

- Describing relationship between variables with a binary outcome
    - Quantitative Explanatory variable:
        - Can use xyplot, especially with jittered data
        - Boxplots 
    - Binary Explanatory variable:
        - 2-by-2 tables: Conditional Proportions, Percentages 
        - Bar chart
        
- Testing Linearity in Simple Logistic Regression (p.472-476)
    - Quantitative: Empirical Logits plot ('binning')
    - Binary: 2 points make a line
    
## Simple -> Multiple Regression

- More explanatory/predictor variables
- Combinations of variable types (quantitative, binary, categorical)
    - Transformed variables
    - Interactions

## Forms of the Model

- Logit form: $log(\frac{\pi}{1-\pi}) = \beta_0+\beta_1 X_1+\beta_2 X_2 + \cdot \cdot \cdot + \beta_k X_k$

- Probability form: $\pi = \frac{e^{\beta_0+\beta_1 X_1+\beta_2 X_2 + \cdot \cdot \cdot + \beta_k X_k}}{1+e^{\beta_0+\beta_1 X_1+\beta_2 X_2 + \cdot \cdot \cdot + \beta_k X_k}}$

If you're asked to write the equation, express it in the logit form (unless the question asks specifically about probability).

## CFAU

- How to *choose* your model?
    - Depends on your goals, your research question, etc.
    - Choosing some components in advance to meet your stated hypotheses

- How do you *fit* the model?
    - Using `R`
    - Estimated models using maximum likelihood estimation

- How do you *assess* the model?
    - Are model assumptions met?
        - Linearity: empirical logit plots
        - Randomness & Independence: Knowing your data well
    - How good is the model?
        - Using -2log(L), nested likelihood ratio / drop-in-deviance test

- How do you *use* the model?
    - If you think the assumptions are met and reasonable for making inferences to the population, you can use the model
    - Statistical significance: p-values from z-test or likelihood ratio test
    - Specifying range of possible estimates: CI based on the z-statistics

## ICU - Can we predict Who survives?

A sample of 200 patients from the ICU. We have information on whether they survived 

  \begin{equation}
    Survived=
    \begin{cases}
      0, & \text{Dead} \\
      1, & \text{Lived}
    \end{cases}
  \end{equation}
  
and explanatory variables about their:

- age (`Age`, quantitative, years), 
- AgeGroup (`AgeGroup`, categorical, 1=young (<50yo), 2=middle (50-69), 3=old(70+))
- sex (`Sex`, indictor/binary, 1=Female, 0=Male)
- Infection (`Infection`, indicator/binary, 1=infection is suspected, 0=no infection)
- Systolic Blood Pressure (`SysBP`, quantitative, mm/Hg), and 
- Pulse (`Pulse`, quantitative, heart rate (beats per minute)),
- Emergency Admission (`Emergency`, indicator, 1=emergency admission, 0=elective admission)


```{r}
require(Stat2Data)
data("ICU")

head(ICU)
str(ICU)
```

```{r, fig.width=2, echo=FALSE}
qplot(x=Age, y=Survive, data=ICU, group=Survive, geom="boxplot")
qplot(x=SysBP, y=Survive, data=ICU, group=Survive, geom="boxplot")
qplot(x=Pulse, y=Survive, data=ICU, group=Survive, geom="boxplot")
tally(~AgeGroup+Survive, data=ICU)
tally(~Sex+Survive, data=ICU)
tally(~Infection+Survive, data=ICU)
tally(~Emergency+Survive, data=ICU)
```

## Multiple Logistic Regression Model
```{r, message=FALSE}
icu_m0 <- glm(Survive ~ Age+Sex+SysBP+Pulse, family=binomial, data=ICU)
summary(icu_m0)
exp(cbind(OR = coef(icu_m0), confint(icu_m0)))
lrtest(icu_m0)
```

<!--- The odds of survival for each additional year of age was .97 times the odds of someone a year younger, on average, in this population, adjusted for blood pressure and pulse. In other words, someone one year older has 3% lower odds of surviving from the ICU than someone a year older, on average, after adjusting for blood pressure and pulse. We are 95% confident that the true odds ratio for age on survival is between 0.95 and 0.99 in the population. --->

1. Write down the logit and probability forms of the model

2. Comment on the effectiveness of each of the predictors in the two-predictor modle.

3. Estimate the probability and odds of an young man surviving the ICU

4. Repeat 3 for a young woman and find the odds ratio

5. Redo both 3&4 for a man and woman aged 70+

6. What happens to the OR as age increases? Will that always be the case?

7. What would Age*Sex do? Explain how the coefficients in the model relate to separate linear models for males and females to predict log(odds) of survival based on age

