---
title: "HW 8 Key"
author: "SDS 291"
date: "April 20, 2020"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r, warning=FALSE, error=FALSE, message=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(mosaic)
library(tidyverse)
library(Stat2Data)
data("Titanic")
```

#10.12 / 10.23

##a

```{r}
Titanic$Sex<-relevel(Titanic$Sex, ref="male")
m1012a<-glm(Survived~Age+Sex, data=Titanic, family=binomial)
summary(m1012a)
```

```{r, echo=FALSE}
m1012a_intercept<-coef(m1012a)[1]
m1012a_age<-coef(m1012a)[2]
m1012a_sexFemale<-coef(m1012a)[3]
```

$log(odds) =$ `r m1012a_intercept` + `r m1012a_age` age + `r m1012a_sexFemale` sexFemale

$\hat\pi = \frac{e^{-1.159839+(-.006352age)+(2.465996smoker)}}{1+e^{-1.159839+(-.006352age)+(2.465996smoker)}}$

- *check: if fit the model and wrote out both forms of the model correctly*
- *check-minus: if didn't fit correctly and only wrote one of the forms of the model*

##b

Age negatively associated with survival - as age increases, the log odds of being surviving decreases, adjusting for sex - and the relationship is no longer statistically significant.

Females have higher, and statistically significantly higher, log odds of survival than men, adjusted for age.

- _check-plus: they include correct and well-stated interpretations of odds ratios_
- _check: any resonable answer here that includes both direction and the statistical significance_
- _check-minus: if they don't talk about the direction *and* statistical significance (i.e., if they only mention statistical significance...)_

##c. 

```{r}
newdata = data.frame(Age=18, Sex="male")
prob_18male<-predict(m1012a,newdata, type="response")
odds_18male<-prob_18male/(1-prob_18male)
```

For a 18 year old man, the probability and odds of surviving the Titanic:

- Probability: `r prob_18male`
- Odds: `r odds_18male`

- _check-plus: if they (correctly) interpret the probability and odds in a sentence_
- _check: correctly calculated both_
- _check-minus: if they miscalculated *or* only included one, not both_

##d.

```{r}
newdata = data.frame(Age=18, Sex="female")
prob_18female<-predict(m1012a,newdata, type="response")
odds_18female<-prob_18female/(1-prob_18female)

OR_18<-odds_18female/odds_18male
```

For a 18 year old female, the probability and odds of surviving the Titanic:

- Probability: `r prob_18female`
- Odds: `r odds_18female`

The Odds Ratio (OR) of the odds of death for a a 18 year old woman who smoked compared to a woman who did not smoke is:

OR: $\frac{Odds_{females}}{Odds_{males}}$ = `r OR_18`

- _check-plus: if they (correctly) interpret the probability and odds in a sentence *and* include the OR in a sentence_
- _check: correctly calculated all: odds, probability, and OR_
- _check-minus: if they miscalculated more than one thing, didn't include the OR, or only included odds but not probability_

##e.

```{r}
newdata = data.frame(Age=50, Sex="male")
prob_50male<-predict(m1012a,newdata, type="response")
odds_50male<-prob_50male/(1-prob_50male)

newdata = data.frame(Age=50, Sex="female")
prob_50female<-predict(m1012a,newdata, type="response")
odds_50female<-prob_50female/(1-prob_50female)

OR_50<-odds_50female/odds_50male
```


| Sex/Outcome| 18yo female | 18yo male | OR | 50yo female | 50yo male | OR |
|--------------|--------------|----------------|------|--------------|----------------|------|
|Probability | `r prob_18female` | `r prob_18male` |   | `r prob_50female` | `r prob_50male` |   |
| Odds | `r odds_18female` | `r odds_18male` |`r OR_18` |  `r odds_50female` | `r odds_50male` | `r OR_50` |

- _check-plus: if they (correctly) interpret the probability and odds in a sentence_
- _check: correctly calculated both odds and probability for both males and females_
- _check-minus: if they miscalculated *or* only included one, not both_

##f.

```{r}
exp(coef(m1012a))
```

The odds ratio of survival for females compared to males is always the same for all ages - this is a reflection of "controlling for age".

Age here is the (linear) slope and females and males just have different intercepts; this is like the parallel slopes model we saw before in linear regression. The slope of the lines are still parallel, so the ratio of the odds for females compared to males is the same at all ages.

- _check-plus: if they give an thorough answer of how and why_
- _check: if they say "yes" it will be the same at all ages without much explanation_
- _check-minus: if they reach the wrong conclusion_

\newpage

# **RECOMMENDED**

No need to grade these
#10.13 / 10.24

##a

```{r}
m1013a<-glm(Survived~Age*Sex, data=Titanic, family=binomial)
summary(m1013a)
```

```{r, echo=FALSE}
m1_intercept<-coef(m1013a)[1]
m1_age<-coef(m1013a)[2]
m1_sexFemale<-coef(m1013a)[3]
m1_sexage<-coef(m1013a)[4]
```

The addition of the interaction term changes the models for Females and Males -- the slope for age now varies by sex.

$log(odds)_{females} =$ `r m1_intercept` + `r m1_age` age + `r m1_sexFemale` smoker + `r m1_sexage` smoker:age 

$log(odds)_{females} =$ (`r m1_intercept`  + `r m1_sexFemale`) + (`r m1_age` + `r m1_sexage`) $\cdot age$

$log(odds)_{females} =$ `r m1_intercept+m1_sexFemale` + `r m1_age + m1_sexage` $\cdot age$

$log(odds)_{males} =$ `r m1_intercept` + `r m1_age` $\cdot age$



```{r, echo=FALSE}
newdata_1 = data.frame(Age=18, Sex="female", `Age:Sexfemale`=18)
prob_18female_int<-predict(m1013a,newdata_1, type="response")
odds_18female_int<-prob_18female_int/(1-prob_18female_int)

newdata_2 = data.frame(Age=18, Sex="male", `Age:Sexfemale`=0)
prob_18male_int<-predict(m1013a,newdata_2, type="response")
odds_18male_int<-prob_18male_int/(1-prob_18male_int)

OR_18_int<-odds_18female_int/prob_18male_int

newdata_3 = data.frame(Age=50, Sex="female", `Age:Sexfemale`=50)
prob_50female_int<-predict(m1013a,newdata_3, type="response")
odds_50female_int<-prob_50female_int/(1-prob_50female_int)

newdata_4 = data.frame(Age=50, Sex="male", `Age:Sexfemale`=0)
prob_50male_int<-predict(m1013a,newdata_4, type="response")
odds_50male_int<-prob_50male_int/(1-prob_50male_int)

OR_50_int<-odds_50female_int/prob_50male_int
```

We now have a model with different intercepts *and* different slopes, so we wouldn't expect the odds ratio between smokers and non-smokers to be constant -- they now vary by age.  We can see from the equations above that the slope for age among smokers is now smaller that it is for non-smokers. What we see below is that as a woman who smokes gets older, the ratio of the odds of death between her and a similar aged nonsmoker gets smaller.

| Smoking/Outcome| 18yo female | 18yo male | OR | 50yo female | 50yo male | OR |
|--------------|--------------|----------------|------|--------------|----------------|------|
|Probability | `r prob_18female_int` | `r prob_18male_int` |   | `r prob_50female_int` | `r prob_50male_int` |   |
| Odds | `r odds_18female_int` | `r odds_18male_int` |`r OR_18_int` |  `r odds_50female_int` | `r odds_50male_int` | `r OR_50_int` |

```{r}
exp(coef(m1013a))
```

_-0.5 if don't illustrate multiple models _
_-0.5 if no explanation of how the interaction term changed the OR_

##b

```{r}
newtitanic<-Titanic %>% filter(!is.na(Age) & !is.na(Sex))
mSex<-glm(Survived~Sex, data=newtitanic, family=binomial)
anova(mSex,m1013a, test = "Chisq")
```

We can reject the null hypothesis that $\hat\beta_2=\hat\beta_4=0$, since the $\chi^2$ statistic is large (and above the critical value of 6) and the p-value is small (p<0.001) and conclude that adjusting for age and letting the relationship between age and survival differ between males and females does a better job of estimating the log(odds) of death than the simple regression model with just sex alone.


_-0.5 if misestimated (it's fine if the order of models are switched and the test statistic is negative) _
_-0.5 if compared to the model from 10.12 with sex and age instead of a different model with just sex_

#10.19 / 10.31
##a

```{r}
newtitanic2<-Titanic %>% filter(PClass!="*") %>% 
  mutate(Surv_fct=as.factor(Survived), 
    PClass2=as.factor(PClass))
newtitanic2$Surv_fct<-relevel(newtitanic2$Surv_fct, ref="1")
gmodels::CrossTable(newtitanic2$PClass, newtitanic2$Surv_fct, 
  prop.r=TRUE, prop.c=FALSE, prop.chisq = FALSE, prop.t = FALSE)
```

We see that the proportion of survival is highest among 1st Class passengers. 2nd and 3rd class passengers have lower odds of survival, likely based on their further distance from the lifeboats.

_-0.5 if probabilities are miscalculated _
_-0.5 if no explanation of the results in the context of the problem_

##b

```{r, warning=FALSE}
gmodels::CrossTable(newtitanic2$PClass, newtitanic2$Surv_fct, 
  prop.r=TRUE, prop.c=FALSE, prop.chisq = FALSE, prop.t = FALSE,
  chisq = TRUE)
```
$H_0$: There is no relationship between PClass and Survival.
$H_A$: There is a relationship between PClass and Survival.

Since the $\chi^2$ test statistic large (172.519) and the p-value is <0.001, we can reject the null hypothesis and conclude that there is a relationship between PClass and Survival.

_-0.5 if no hypotheses are mentioned _

##c
```{r}
m1019c<-glm(Survived~PClass2, data=newtitanic2, family=binomial)
summary(m1019c)
```

```{r}
exp(coef(m1019c))
```

The odds of surviving for a passenger in 1st class is 1.5 to 1. These odds are significantly greater than 0.

The odds of surviving for a passenger in 2nd class is 0.5 times (i.e., lower) than for 1st Class passengers. This difference was statistically significant, as evidenced by the z statistic>1.96 and the p-value<0.05.

The odds of surviving for a passenger in 3rd class is 0.16 times (i.e., lower) than for 1st Class passengers.

_-0.5 if miscalculated (if PClass* is the reference, and all three PClass variables are in the model) _
_-0.5 if no interpretation_

##d

```{r}
m1019c_1st<-coef(m1019c)[1]
m1019c_2nd<-coef(m1019c)[2]
m1019c_3rd<-coef(m1019c)[3]


newdata_1st = data.frame(PClass2="1st")
prob_1st<-predict(m1019c,newdata_1st, type="response")
odds_1st<-prob_1st/(1-prob_1st)

newdata_2nd = data.frame(PClass2="2nd")
prob_2nd<-predict(m1019c,newdata_2nd, type="response")
odds_2nd<-prob_2nd/(1-prob_2nd)

newdata_3rd = data.frame(PClass2="3rd")
prob_3rd<-predict(m1019c,newdata_3rd, type="response")
odds_3rd<-prob_3rd/(1-prob_3rd)

OR_2nd<-odds_2nd/odds_1st

OR_3rd<-odds_3rd/odds_1st
```



| Smoking/Outcome| 1st Class | 2nd Class | OR | 3rd Class |  OR |
|--------------|--------------|----------------|------|--------------|------|
|Probability | `r prob_1st` | `r prob_2nd` |   | `r prob_3rd` |    |
| Odds | `r odds_1st` | `r odds_2nd` |`r OR_2nd` |  `r odds_3rd` | `r OR_3rd` |



We get the same fitted probabilities from the model than we do from the table in #9.

```{r, warning=FALSE}
gmodels::CrossTable(newtitanic2$PClass, newtitanic2$Surv_fct, 
  prop.r=TRUE, prop.c=FALSE, prop.chisq = FALSE, prop.t = FALSE,
  chisq = TRUE)
```

_-0.5 if miscalculated if they don't match _

##e

```{r}
anova(m1019c, test="Chisq")
```

Yes, it does match the answer from above.

_-0.5 if they don't calculate this correctly_