---
title: "Take Home Exam 2"
author: "SDS 291, Spring 2020"
date: "Due: 05/08/2020"
output:
  html_document:
    css: "~/Dropbox/3_Smith/SDS291/sds291_site/static/img/labs.css"
    highlight: textmate
    theme: journal
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, error=FALSE, message=FALSE, warning=FALSE}
data<-read.csv(url("http://sds291.netlify.app/exam/291-2ndExam.csv"))
library(tidyverse)
```


# Low birthweight and Number of Hospital Visits (`visits`)

*1. Generate a visual of the relationship between your explanatory and response variable and interpret what you see. Do not include the regression line.*

```{r, error=FALSE, warning=FALSE, message=FALSE}
qplot(x=visits, y=lowCode, data=data) + geom_jitter()
ggplot(data, aes(x=visits, y=lowbirthweight)) + geom_boxplot()
```

*2. Use R to estimate a simple logistic regression model and write the fitted regression equation in the logit form. You should have one equation expressed in two different ways, one with the fitted values (i.e., numbers) and one way without the numbers.*

```{r}
model1<-glm(lowCode~visits, data=data, family=binomial)
summary(model1)
```

*3. Calculate and interpret the odds ratio of the slope in a sentence, in the context of this problem, and conduct a hypothesis test of the significance of this odds ratio.*

```{r}
exp(coef(model1)[2])
```


*4. Calculate the odds and probability of having a child born at low birthweight for birth parent who had 3, 4, 14, and 15 hospital visits.*

```{r}
visit_nums<-c(3,4,14,15)
logodds<-coef(model1)[1]+ (coef(model1)[2]*visit_nums)
m1_data<-as.data.frame(cbind(visit_nums,logodds))
m1_data <- m1_data %>%
  mutate(odds=exp(logodds),
        prob = odds/(1+odds))

m1_data
```

*5. Interpret the calculated odds and probabilities of low birthweight for 3 and 4 visits in complete sentences in the context of the problem.*

*6. Connect these odds for all 4 of these values of visits to the hospital to the odds ratio you calculated in Exercise 3.*

```{r}
OR_34<-m1_data$odds[2]/m1_data$odds[1]
OR_1415<-m1_data$odds[4]/m1_data$odds[3]

OR_34
OR_1415
```


# Low birtweight and Marital Status (`marital`)

Marital status can be a reflection of socioeconomic status - dual earning households, potentially shared access to health insurance afforded to married couples - that may be related to birth outcomes. The following 4 questions refer to the variable in this dataset that reflects marital status (`marital`). [**Note:** do not change the `marital` variable -- use it exactly as is.]

```{r, echo=FALSE}
gmodels::CrossTable(data$marital,data$lowbirthweight, prop.r = FALSE, prop.c=FALSE,prop.t = FALSE, prop.chisq = FALSE)
```

*7. Use the information from the two-way table above to calculate the odds and probabilities of having a child who is low birthweight for married and unmarried parents.*

```{r}
odds_married <- 60/320
odds_unmarried <- 48/562

odds_married
odds_unmarried

```

*8. Fit a logistic regression model to estimate the relationship between low birthweight and marital status. Use the estimates from this fitted model to calculate the odds and probabilties of having a child who is low birthweight for married and unmarried parents.*

```{r}
model2<-glm(lowCode~marital, data=data, family=binomial)
summary(model2)

married<-c(0,1)
logodds<-coef(model2)[1]+ (coef(model2)[2]*married)
m2_data<-as.data.frame(cbind(married,logodds))
m2_data <- m2_data %>%
  mutate(odds=exp(logodds),
        prob = odds/(1+odds))

m2_data
```

*9. Calculate the odds ratio from both the two-way table and from the regression output, and interpret this odds ratio in the context of this problem.*

```{r}
OR_2way<-odds_unmarried/odds_married
OR_2way

exp(coef(model2)[2])
```

*10. Calculate the 95% Confidence Interval for the odds ratio and interpret this value in a sentence in the context of the problem.*

```{r}
MarriedORCI<-exp(cbind(coef(model2), confint(model2)))
MarriedORCI
```

# Extending Each Model, Revisiting Gestational Duration

You may remember that gestational duration (`weeks`) was important when you considered birthweight as a quantitative variable in linear regression. You want to use that again, and add it to the models that you have already built with `visits` and `marital`. The following 3 questions refer to the variables you have already used, plus also `weeks` that you used in the first exam.

*11. Fit a multiple logistic regression model of low birthweight with `weeks`, `visits`, and `marital` as explanatory variables. *

```{r}
model3<-glm(lowCode~marital+visits+weeks, data=data, family=binomial)
summary(model3)
```

*12. Interpret the odds ratio for each explanatory variable in the context of this problem, and comment on the statistical significance of each variable.*

*13. Is this model better than the simple logistic regression model you fit in Exercise 2? Conduct a test that compares the models, and explain what you conclude in the context of this problem (i.e., what does this test tell us about these particular relationships)?*

```{r}
anova(model1,model3)
lmtest::lrtest(model3,model1)
```


# Communicating Take-Away Message

*14. Imagine your supervisor is preparing to invest in a prevention campaign to reduce low birthweight. Based on your models, what do you think she should focus on? Summarize in 2-3 sentences what you found through these models about risk factors for low birthweight for your supervisor. Use this as an opportunity to illustrate you understand statistical models, but without overly technical details.*
