---
title: "Simple Linear Regression"
author: "YOUR NAME"
date:   "TODAY'S DATE"
output: pdf_document
---

```{r include=FALSE}
# Some customization.  You can alter or delete as desired (if you know what you are doing).

# This changes the default colors in lattice plots.
#trellis.par.set(theme=theme.mosaic())  

# knitr settings to control how R chunks work.
require(knitr)
opts_chunk$set(
  tidy=FALSE,     # display code as typed
  size="small"    # slightly smaller font for code
)
```


# Bring in the packages and the data
```{r, message=FALSE, error=FALSE}
require(tidyverse)  # loading the data management suite of tools
require(ggplot2)    # the plotting functions for making graphs  
data("anscombe")    # the data we will be using - this brings it into R memory for you to use
```

## Choose - Scatterplot

As a first step to choosing an appropriate model, produce a scatterplot of the variables of interest:

```{r}
ggplot(data = anscombe, aes(x = x1, y = y1))  + #identifying the dataset and variable names
    geom_point()                              # Plot the scatterplot 
```

`qplot`, or "quick" plot, is part of the ggplot/tidyverse package. This is a good way to get basic descriptive visualizations - histograms, boxplots, scatterplots, etc.  Since you specify an x and a y value, the default is to provide a scatterplot; for other specifications, you might need more code.

1. Describe the relationship seen in the scatterplot above.  Think about each of the following 5 dimensions of scatterplots in your answer:

- *Direction* (positive/negative)
- *Linearity* (is there a curve or shape?)
- *Unusual Observations* (outliers or points that don't follow the same pattern?)
- *Strength* of relationship (are dots close together along a line or more spread out?)
- *Magnitude* of the relationship (is the slope steep or shallow?)




1. Write the equation for the **hypothetical model** we would like to fit?



## Fit - Estimating the Model
Use the `lm()` function to fit this model. Remember to save the regression model with a name (here, `m1`). Then use the `summary()` function to view the details of the model output.

```{r}

#fit the regression model
m1<-lm(y1 ~ x1 , data = anscombe)
summary(m1)

#visualize the regression model
ggplot(data = anscombe, aes(x = x1, y = y1))  +
    geom_point() +           # Plot the scatterplot 
    geom_smooth(method=lm,   # Add linear regression line
                se=FALSE)    # Don't add shaded confidence region
```


1. Write the **fitted** regression equation.

1. Interpret the values of the Intercept and of x1, respectively, in a sentence.

1. What value of y1 would you expect if x1 was 6.5?  [Note: you can do the math from the equation above or get R to do the prediction for you - see below]

```{r, error=FALSE, message=FALSE}
library(mosaic)
ans_predict<-makeFun(m1)
ans_predict(x1=6.5)
```

# Assess - Is a linear regression model a good fit for these data?

## Residuals and Predictions

To bring the residuals into a dataframe to use and for `ggplot` to plot, we'll use the `broom` package (think of a broom sweeping up the messy regression results into a pile) to save the actual/observed and fitted/predicted values and the residuals together into a dataframe we will call `m1_data`.

```{r, error=FALSE, message=FALSE, warning=FALSE}
library(broom)
m1_data<-augment(m1)
#View(m1_data)
```

1. Find the residual for where x1=8.

## Regression Assumptions

1. Generate relevant plots to evaluate regression assumptions.  

Note, there are a few ways to do this. One uses "base" R plots and the other uses `ggplot`; they get you the same information, but `ggplot` is more aesthetically pleasing, easier to modify, more flexible in terms of format, etc.
```{r}
plot(m1, which=c(1:2))
```

# Fitted-Residuals Plot

```{r fitted-residual plot}
ggplot(data = m1_data, aes(x = .fitted, y = .resid)) + geom_point()
```

## Q-Q Plot

```{r qqplot, fig.height=4}
ggplot(data = m1_data, aes(sample = .resid)) + geom_qq()
```

## Histogram of Residuals

```{r histogram of residuals, fig.height=4}
ggplot(data = m1_data, aes(x = .resid)) + geom_histogram()
```

1. Evaluate the regression plots with respect to the regression assumptions.

# Hypothesis testing

1. Test the hypothesis that x1 has a linear relationship with y1. Report and interpret the p-value for the test. [Note: be sure the state the null and alternative hypotheses for the test]

1. Find and interpret (in a complete sentence) a 95% confidence interval for the slope.

```{r}
confint(m1)
```

# More practice

## Residual Squared Error
1. Practice calculating the RSE in R.  Explain line by line what's happening in the code below.

# Regression/residual standard error

$\hat{\sigma_{\epsilon}} = \sqrt{\frac{\sum(y - \bar{y}) ^2}{n-2}} = \sqrt{\frac{SSE}{n-2}}$

```{r SSE RSE}
SSE <- sum((m1_data$.resid)^2)
SSE
RSE<-sqrt(SSE/9)
RSE
```

1. Do you get the same answers as what's reported in the linear model results and ANOVA table (we'll talk more about the ANOVA table next class -- just see if you can find the same number for now)

```{r linear model and anova table}
summary(m1)
anova(m1)
```

1. Interpret the residual standard error in a sentence.

