---
title: "Sections 1.2-3 - Assssing Conditions for a Simple Linear Model"
author: "Stat2"
date: "July 3, 2018"
output:
  pdf_document: default
---

Load needed packages.
```{r message=FALSE, warning=FALSE}
library(Stat2Data)
library(e1071) #used to create a different type of probability plot with probplot() funtion
```

EXAMPLE 1.5 Accord prices (continued)

Get the dataframe with **AccordPrice** data.
```{r}
data(AccordPrice)
str(AccordPrice)
```

Find the least-squares regression line.
```{r}
regmodel=lm(Price~Mileage, data=AccordPrice)
summary(regmodel)
```

The fitted values and many other statistics are stored in the regmodel object.
```{r}
fits=regmodel$fitted.values
fits
```

The residuals are also stored in the regmodel object.  These will be very useful in checking conditions.
```{r}
resids=regmodel$residuals   #store the residual values in a varaible named "resids"
resids
```

EXAMPLE 1.5 Accord prices (continued) 

Compute the standard error of regression (by formula)
```{r}
n=length(AccordPrice$Mileage)
SSE=sum(resids^2)                      #Sum of squared errors
Se=sqrt(SSE/(n-2))
Se
```
Note that you can find this value in the regression summary, labeled as "Resiudal standard error."

The code below will generate plots showing different sorts of results for residual versus fits and normal quantile plots. 

FIGURE 1.5 Residuals versus fitted values plot when linearity and constant variance conditions hold

```{r}
## ORIGINAL
set.seed(11)    # for replcating the graphs
residual=rnorm(100)
YHat=runif(100,20,50)
df_het<-as.data.frame(cbind(YHat,residual))
plot(df_het$residual~df_het$YHat,ylim=c(-3,3),cex.lab=1.5,cex.axis=1.5,cex=1.5)
abline(0,0,col="darkblue")


## BENS VERSION

library(tidyverse)
library(ggridges)

#THIS WORKS
res1 <- rnorm(1000)
YHat1 <- runif(1000,20,50)
residuals_1<-as.data.frame(cbind(YHat1,res1))
residuals_1$YHat2 <- cut(residuals_1$YHat1, seq(20, 50, 2), include.lowest = TRUE)

ggplot(residuals_1,aes(x = res1, y = YHat2)) + 
  geom_density_ridges() + theme(axis.text.x = element_text(angle=45)) +
  xlab("Residuals") + ylab("Fitted Values (estimated value of y-hat)") + 
  #theme(axis.text.x = element_text(angle=45)) +
  scale_y_discrete(labels= c("20", "22", "24", "26", "28",
                             "30", "32", "34", "36", "38",
                             "40", "42", "44", "46", "48")) +
  coord_flip()


constant_1<-ggplot(residuals_1,aes(x = res1, y = YHat2)) + 
  geom_density_ridges()  +
  xlab("Residuals") + ylab("Fitted Values (estimated value of y-hat)") + 
  #theme(axis.text.x = element_text(angle=45)) +
  scale_y_discrete(labels= c("20", "22", "24", "26", "28",
                             "30", "32", "34", "36", "38",
                             "40", "42", "44", "46", "48")) +
  geom_vline(xintercept = c(-1,1), color="black", size=.3 , linetype = "dashed") + 
  geom_vline(xintercept = c(-2,2), color="black", size=.3 )

constant_4850<-
  constant_1 + 
  geom_vline(xintercept = c(-1,1), color="red", size=.3, linetype="dashed") +
  geom_vline(xintercept = c(-2,2), color="red", size=.3)

constant_2022<-
  constant_1 +
  geom_vline(xintercept = c(-1,1), color="blue", size=.3, linetype="dashed") +
  geom_vline(xintercept = c(-2,2), color="blue", size=.3)



```

FIGURE 1.6 Residuals versus fitted values plots illustrating problems with conditions

FIGURE 1.6a - nonlinearity
```{r}
YHat=runif(100,20,50)
newresidual=rnorm(100)+0.05*(YHat-32)^2
residual=(newresidual-mean(newresidual))/sd(newresidual)
plot(residual~YHat,cex.lab=1.5,cex.axis=1.5,cex=1.5)
abline(0,0,col="darkblue")


#THIS WORKS
set.seed(11)
YHat_nl <- runif(1000,20,50)
newres_nl <- rnorm(1000)+0.05*(YHat_nl-32)^2
res_nl=(newres_nl-mean(newres_nl))/sd(newres_nl)
residuals_nl<-as.data.frame(cbind(YHat_nl,res_nl))


nl_scatter <- qplot(x=YHat_nl, y=res_nl, data=residuals_nl) + theme_bw() +
  xlab("Residuals") + ylab("Fitted Values (estimated value of y-hat)") +
  geom_hline(yintercept = 0, color="black", size=.3 , linetype = "dashed")
nl_scatter

residuals_nl$YHat_nl2 <- cut(residuals_nl$YHat_nl, seq(20, 50, 2), include.lowest = TRUE)

nl_ridge <- ggplot(residuals_nl,aes(x = res_nl, y = YHat_nl2)) + 
  geom_density_ridges() +coord_flip() + theme_bw() + 
  #geom_vline(xintercept = 0, color="blue", size=.5 ) + 
  xlab("Residuals") + ylab("Fitted Values (estimated value of y-hat)") + 
  #theme(axis.text.x = element_text(angle=45)) +
  scale_y_discrete(labels= c("20", "22", "24", "26", "28",
                             "30", "32", "34", "36", "38",
                             "40", "42", "44", "46", "48"))

```

FIGURE 1.6b - nonconstant variability
```{r}
YHat=runif(100,20,50)
newresidual=(YHat-19)*rnorm(100)
residual=(newresidual-mean(newresidual))/sd(newresidual)
plot(residual~YHat,cex.lab=1.5,cex.axis=1.5,cex=1.5,ylim=c(-3,3))
abline(0,0,col="darkblue")


YHat_nc <- runif(10000,20,50)
newres_nc <- rnorm(10000)*(YHat_nc-19)
res_nc=(newres_nc-mean(newres_nc))/sd(newres_nc)
residuals_nc<-as.data.frame(cbind(YHat_nc,res_nc))
qplot(x=YHat_nc, y=res_nc, data=residuals_nc)

residuals_nc$YHat_nc2 <- cut(residuals_nc$YHat_nc, seq(20, 50, 2), include.lowest = TRUE)

ggplot(residuals_nc,aes(x = res_nc, y = YHat_nc2)) + 
  geom_density_ridges() + theme_bw() +
  #geom_vline(xintercept = 0, color="blue", size=.5 ) + 
  xlab("Residuals") + ylab("Fitted Values (estimated value of y-hat)") + 
  #theme(axis.text.x = element_text(angle=45)) +
  scale_y_discrete(labels= c("20", "22", "24", "26", "28",
                             "30", "32", "34", "36", "38",
                             "40", "42", "44", "46", "48")) +
  coord_flip() 


# residuals_nc %>%
#   filter(YHat_nc2=="[20,22]") %>%
#   summarise(mean_1stgrp = mean(res_nc),
#          sd_1stgrp=sd(res_nc)) %>%
#   mutate(l1sd=mean_1stgrp - sd_1stgrp,
#          u1sd=mean_1stgrp + sd_1stgrp,
#          l2sd=mean_1stgrp - (2*sd_1stgrp),
#          u2sd=mean_1stgrp + (2*sd_1stgrp),
#     )
# 
# residuals_nc %>%
#     filter(YHat_nc2=="(48,50]") %>%
#   summarise(mean_1stgrp = mean(res_nc),
#          sd_1stgrp=sd(res_nc)) %>%
#   mutate(l1sd=mean_1stgrp - sd_1stgrp,
#          u1sd=mean_1stgrp + sd_1stgrp,
#          l2sd=mean_1stgrp - (2*sd_1stgrp),
#          u2sd=mean_1stgrp + (2*sd_1stgrp),
#     )

ggplot(residuals_nc,aes(x = res_nc, y = YHat_nc2)) + 
  geom_density_ridges() + theme_bw() +
  geom_vline(xintercept = c(-1,1), color="black", size=.3 , linetype = "dashed") + 
  geom_vline(xintercept = c(-2,2), color="black", size=.3 ) + 
  #geom_vline(xintercept = c(-1.6988,1.622518), color="red", size=.3, linetype="dashed") +
  #geom_vline(xintercept = c(-3.359519,3.283197), color="red", size=.3) +
  #geom_vline(xintercept = c(-0.110156,0.1220279), color="blue", size=.3, linetype="dashed") +
  #geom_vline(xintercept = c(-0.226248,0.2381198), color="blue", size=.3) +
  xlab("Residuals") + ylab("Fitted Values (estimated value of y-hat)") + 
  #theme(axis.text.x = element_text(angle=45)) +
  scale_y_discrete(labels= c("20", "22", "24", "26", "28",
                             "30", "32", "34", "36", "38",
                             "40", "42", "44", "46", "48"))




residuals_nc %>%
  filter(YHat_nc2=="[20,22]") %>%
  summarise(mean_1stgrp = mean(res_nc),
         sd_1stgrp=sd(res_nc)) %>%
  mutate(l1sd=mean_1stgrp - sd_1stgrp,
         u1sd=mean_1stgrp + sd_1stgrp,
         l2sd=mean_1stgrp - (2*sd_1stgrp),
         u2sd=mean_1stgrp + (2*sd_1stgrp),
    )

residuals_nc %>%
    filter(YHat_nc2=="(48,50]") %>%
  summarise(mean_1stgrp = mean(res_nc),
         sd_1stgrp=sd(res_nc)) %>%
  mutate(l1sd=mean_1stgrp - sd_1stgrp,
         u1sd=mean_1stgrp + sd_1stgrp,
         l2sd=mean_1stgrp - (2*sd_1stgrp),
         u2sd=mean_1stgrp + (2*sd_1stgrp),
    )

```

FIGURE 1.6c - Problem with linearity and constant variance
This example uses the Perch data, then transforms.
```{r}
data(Perch)
model1=lm(Weight~Length,data=Perch)
YHat=(model1$fitted+400)/50
newresidual=model1$residual
residual=(newresidual-mean(newresidual))/sd(newresidual)
plot(residual~YHat,cex.lab=1.5,cex.axis=1.5,cex=1.5,ylim=c(-3,3))
abline(0,0,col="darkblue")
```

FIGURE 1.7 Examples of normal quantile plots

FIGURE 1.7a  A "good" example
```{r}
residual=rnorm(100)
qqnorm(residual,ylab="Residual",cex.lab=1.5,cex.axis=1.5,cex=1.5,main=NULL)
qqline(residual,col="darkblue")
```

FIGURE 1.7b  Skewed residuals
```{r}
residual=rexp(100)
qqnorm(residual,ylab="Residual",cex.lab=1.5,cex.axis=1.5,cex=1.5,main=NULL)
qqline(residual,col="darkblue")
```

FIGURE 1.7c  Long tails.

This example uses MetroHealth83 data from first edition.
```{r}
data("MetroHealth83")
model2=lm(NumMDs~NumHospitals,data=MetroHealth83)
newresidual=model2$residual
residual=(newresidual-mean(newresidual))/sd(newresidual)
qqnorm(residual,ylab="Residual",cex.lab=1.5,cex.axis=1.5,cex=1.5,main=NULL)
qqline(residual,col="darkblue")
```

FIGURE 1.8 Examples of normal probability plots



Figure 1.8a Shows normal probability plots from Minitab, which are less common in R.  Most R applications tend to use the QQ plot instead. There is a probplot function in the e1071 pacakge, which looks simialr to the Minitab plots.

```{r}
residual=rnorm(100)
probplot(residual)
```

FIGURE 1.8b A Minitab-like normal plot with problems
```{r}
#using the residuals from Figure 1.7c
probplot(model2$residuals)
```

EXAMPLE 1.6 Accord prices-checking conditions

Add the regression line to the scatterplot.
```{r}
plot(Price~Mileage, data=AccordPrice)
abline(regmodel)
```

FIGURE 1.9 Plot of Accord residuals versus fitted values

Scatterplot of residuals versus fits (with horizontal line at zero for reference)
```{r}
plot(resids~fits)
abline(0,0)
```

FIGURE 1.10 Histogram of Accord residuals

```{r}
hist(resids)
```

FIGURE 1.11 Normal quantile plot of residuals for Accord data

```{r}
qqnorm(resids)
qqline(resids)
```

We can also use the plot command on a regression model to generate several diagnostic plots

```{r}
plot(regmodel)
```

----------------------------------------------------------
Alternative Solutions

Note: Several of the residual plots from plot(regmodel) use standardized residuals which aren't introduced until Topic 4.3. These are easy to find directly from a model.  They can be substituted for ordinary residuals in any of the residual plots.
```{r}
stdres=rstandard(regmodel)
stdres
studres=rstudent(regmodel)
studres
```

Plot standardized residuals, instead of residuals, against fitted values.
```{r}
plot(stdres~regmodel$fitted.values,ylim=c(-3,3),cex.lab=1.5,cex.axis=1.5,cex=1.5)
abline(0,0,col="darkblue")
```

