---
title: "Day 10 - Categorical Variables and Nested F Tests"
author: "SDS 291"
date: "February 26, 2020"
output:
  html_document:
    css: "~/Dropbox/3_Smith/SDS291/sds291_site/static/img/labs.css"
    highlight: textmate
    theme: journal
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy=FALSE)
```


## Birthweight and parental smoking
Using data from the mosaic package. Birth weight, date, and gestational period collected as part of the Child Health and Development Studies from Oakland, CA in 1961 and 1962; we're working with a sample of 1,263 babies and their parents. The study is [still ongoing](http://www.chdstudies.org/index.php), now following its 3rd generation.

- Response variable
    - `wt`: birth weight (in ounces) 
    
- Explanatory Variable(s)
    - `smoke`: smoke does mother smoke? 0=never, 1=smokes now, 2=until current pregnancy, 3=once did, not now
    - `age`: mother's age in years at termination of pregnancy

Bring in the data

```{r,  message=FALSE, echo=TRUE}
require(mosaic)
require(tidyverse)
require(magrittr)
data("Gestation")
Gestation<-Gestation %>% filter(!is.na(smoke), !is.na(wt))
```

Now fit the model: $$bithweight=\beta_0+\beta_1 smoke+ \epsilon$$
```{r,  message=FALSE, echo=TRUE}
m_quant<-lm(wt~smoke, data=Gestation)
summary(m_quant)
```

1. How do you interpret the coefficient for `smoke`?



## What if smoking status were a categorical variable?

###R uses `factor` variables to indicate categorical variables. 

You could make a new variable `smoke_factor` with the same values as `smoke` but where `R` knows the variable should be formatted as a factor.
```{r,  echo=TRUE, message=FALSE}
Gestation<-Gestation %>%
  mutate(smoke_factor=as.factor(smoke))
m_cat1<-lm(wt~smoke_factor, data=Gestation)
summary(m_cat1)
```

1. What is the average expected birthweight of a child whose mother never smoked?

 <!-- **Answer**: The average expected birthweight of a child whose mother never smoked is 122.78 ounces.  -->

1. Interpret the coefficient for `smoke_factor1`.

 <!-- **Answer**: The coefficient for `smoke_factor1` reflects the difference in the average expected birthweight of a child whose mother is a current smoker to a child whose mother never smoked. Specifically, a child born to a mother who is a current smoker will weigh 8.67 ounces lighter at birth than a child born to a mother who never smoked, on average in this population. This -8.67 ounce difference in birthweight between current and never smoker mothers is statisticially significant from 0 since the t-statistic (-7.8) is below the critical value of 1.96, and the p-value (9.53e-15) is <0.05.  -->

1. Interpret the coefficient for `smoke_factor2`.

 <!-- **Answer**: The coefficient for `smoke_factor2` reflects the difference in the average expected birthweight of a child whose mother who smoked until pregnancy to a mother who never smoked. A child of a woman who smoked prior to pregnancy weighed, on average, .3066 ounces more than a child of a never smoker mother in this population. This difference was not statistically signigicantly different from 0, since the t-statistic (0.156) was <1.96 and the p-value (0.876) was >0.05.  -->

It's hard to keep the values of these levels straight. Especially if you had multiple factor variables. Instead, you might make the factor levels be something more conceptually understandable.

```{r,  echo=TRUE}
Gestation<-Gestation %>%
  mutate(smoke_cat=as.factor(if_else(smoke==0,"never smoker",
                             if_else(smoke==1,"current smoker",
                             if_else(smoke==2,"pre-pregnancy smoker",
                             if_else(smoke==3,"other former smoker","NA"))))
                   )
        )
tally(~smoke_cat, data=Gestation)
smoke_factor_labels<-lm(wt~I(smoke_cat), data=Gestation)
summary(smoke_factor_labels)
```

1. What is the average expected birthweight of a child whose mother never smoked?

<!-- **Answer**: The average expected birthweight of a child whose mother never -->
<!-- smoked was 122.78 ounces (114.109+8.668). -->

1. What is the average expected birthweight of a child whose mother was a current smoker?

 <!-- **Answer**: The average expected birthweight of a child whose mother was a current smoker was 114.11 ounces.  -->

## How to get the right reference group?

You can tell `R` what you want the reference group to be with `relevel()` function.

```{r,  echo=TRUE}
Gestation$smoke_cat<-relevel(Gestation$smoke_cat, ref = "never smoker")
```

But *the better option* is to make multiple indicator variables - one for each of the categories.
```{r,  echo=TRUE, message=FALSE}
Gestation<-Gestation %>%
  mutate(
    smoke_nev=if_else(smoke==0,1,0),
    smoke_cur=if_else(smoke==1,1,0),
    smoke_pre=if_else(smoke==2,1,0),
    smoke_fmr=if_else(smoke==3,1,0)
  )
tally(c("smoke_nev", "smoke_cur", "smoke_pre", "smoke_fmr"), data=Gestation)
smoke_indicators<-lm(wt~smoke_nev+smoke_cur+smoke_pre+smoke_fmr, data=Gestation)
summary(smoke_indicators)
```


## What does the ANOVA table tell us?
```{r,  echo=TRUE, message=FALSE}
anova(smoke_indicators)
```

1. What's the MSModel?

 <!-- The MSModel is 7879.333 ((23399.5+.1+238.4)/3), which has an F-statistic of 25.19 on 3 and 1222 DF. This model explains a statistically signficant more of the variance in birthweight than a constant model with no variables, as evidenced by the small p-value (8.196e-16) which is <0.05.  -->



## What about a more parsimonious model?

Webster's Dictionary defines parsimony as "the quality of being careful with money or resources." Essentially, we use parsimonious as an adjective to describe a simpler model; we're being careful with our degrees of freedom and number of coefficients in the model, and would prefer to "spend" fewer of them. A simpler model is generally better -- it's easier to explain.

```{r,  echo=TRUE, message=FALSE}
Gestation<-Gestation %>%
  mutate(smoke_cur_d=as.factor(if_else(smoke==1,"Smoker","NonSmoker")))
smoke_cur_d<-lm(wt~smoke_cur_d, data=Gestation)
summary(smoke_cur_d)
```


1. How does this change our interpretation of `smoke_cur_d`?

 <!-- **Answer**: The reference group is now not-current-smokers, a combination of never smokers and former smokers (both those pre-pregnancy and who had quit longer before pregnancy). A child of a mother who currently smokes will be, on average, 8.94 ounces lighter than a mother who does not currently smoke in this population.  -->

1. Do we need the 4 category variable or is the dichotomous/binary variable enough?

```{r,  echo=TRUE, message=FALSE}
anova(smoke_cur_d,smoke_indicators)
```

<!-- **Answer**: We fail to reject the null hypothesis that the coefficients for smoke_pre and smoke_fmr both equal 0. We can conclude that, since these extra coefficients in the model were together not different from 0, that the nested, parsimonious model of a binary variable (current vs. non-current smokers) is sufficient)  -->

What we want to do is to test whether the model with four categories of smoking is _better_ than the a binary definition of smoking (current vs. not current, which includes never, former, and smokers until pregnancy). Essentially, is the right plot with a binary explanatory variable better than the left of a model with four categories of `smoke`:

```{r, echo=TRUE, warning=FALSE, message=FALSE, fig.height=4}
library(moderndive)
qplot(y=wt, x=age, color=smoke_cat, data=Gestation)+geom_parallel_slopes(se=FALSE)
wt4cat_temp<-lm(wt~smoke_cat+age, data=Gestation)
```


```{r, echo=TRUE, warning=FALSE, message=FALSE, fig.height=4}
qplot(y=wt, x=age, color=smoke_cur_d, data=Gestation)+geom_parallel_slopes(se=FALSE)
smoke_cur_dage<-lm(wt~smoke_cur_d+age, data=Gestation)
```


We are essentially testing the hypothesis that former smokers and quit at pregnancy are the same as never smokers vs. the alternative that one of them is different. 

$$H_0: \beta_\text{former} = \beta_\text{quit-at-pregancy} = 0 $$
$$H_A: \beta_i \neq 0 $$

##### Nested Model
```{r}
anova(smoke_cur_dage)
```

##### Full Model
```{r}
anova(wt4cat_temp)
```

##### Nested F Test
```{r}
anova(smoke_cur_dage, wt4cat_temp)
```

The nested-F test is $$Nested F = \frac{\frac{SSM_\text{full}-SSM_\text{nested}}{\text{Number of predictors}}}{\frac{SSE_\text{Full}}{n-k-1}}$$

In this case, we have $Nested F = \frac{\frac{23632-23393}{2}}{\frac{382224}{1219}} = \frac{\frac{230.14}{2}}{313.5} = \frac{115.07}{313.5} = 0.367$

We fail to reject the null hypothesis (the F statistic is very small, <1, and the p-value is very large and above our standard threshold 0.69>0.05) and conclude that we don't have evidence that former and quit-at-pregnancy aren't the same as each other.  Thus, a model with just non-smokers vs. smokers does just as well as a model with all four categories (current vs. never, former, quit-at-pregnancy). 