---
title: "SLR/2D into MLR/3D"
author: "SDS 291, Fall 2020"
date: "2/17/2020"
output:
  html_document:
    css: "~/Dropbox/3_Smith/SDS291/sds291_site/static/img/labs.css"
    highlight: textmate
    theme: journal
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(Stat2Data)
data("AccordPrice")
library(tidyverse)
```

Using the `AccordPrice` Data, let's work with the book's example that we know well at this point.

## Simple Linear Regression

We're fitting a regression model of Mileage and Price.

```{r 2d}
qplot(x=Mileage, y=Price, data=AccordPrice) + geom_smooth(method="lm", se=FALSE) + theme_bw()
```

```{r, echo=FALSE}
m0<-lm(Price~Mileage, data=AccordPrice)
summary(m0)
```

## Now Multiple Regression - 3D

You may want to adjust for the car's age. Lots of miles on a newer car means something different than lots of miles on an older car (the condition of the car overall has more wear and is more outdated).

```{r 3d, echo=FALSE, warning=FALSE, message=FALSE, error=FALSE}
# Note: You'll *probably* need to install the `rgl` package once before running this.
# This does not work on the server -- the RGL package will not install there. 
# Find a desktop version (like the computers in SR 301 or in other computer labs) to test this out

#install.packages("rgl")

library(rgl)
open3d()
Price<-AccordPrice$Price
Mileage<-AccordPrice$Mileage
Age<-AccordPrice$Age
fit<-lm(Price~Mileage+Age)
plot3d(Mileage, Age, Price, type = "s", col = "red", size = .5)

coefs <- coef(fit)
a <- coefs["Mileage"]
b <- coefs["Age"]
c <- -1
d <- coefs["(Intercept)"]
planes3d(a, b, c, d, alpha = 0.5)

rglwidget(width = 1000, height = 1000, reuse = FALSE)
```


```{r, echo=FALSE}
summary(fit)
```

# Connecting the 2D with 3D

1. Can you move the 3D box around to the point that it depicts (something very close to) the 2D / simple relationship between Mileage and Price?

1. How does the quantitative value of the slope parameter(s) correspond to the visual? From the numeric output, would you have expected a steep or a shallow slope? Is that what you see visually?

1. How is the numerical concept of "controlling for" expressed visually?

1. What will the residuals look like in the 3D model?
