---
title: "Chi-Square test by hand"
author: "SDS 291"
date: "4/4/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

From p.465 in the book:

| Marital Status  | Boy Child | Girl Child | Total |
|------------|-----------|--------------|----------|
| Married | 176 | 148 | 324 |
| Did Not Marry | 134 | 142 | 276 |
| Total | 310 | 290 | 600 |

$H_0:$ Single Mothers Marrying and Child Gender are independent

$H_A:$ Single Mothers Marrying and Child Gender are related/dependent 

## Calculating the Expected Values for Each Cell

$E_{ij} = \frac{Row Total \cdot Column Total}{n}$

| Marital Status  | Boy Child | Girl Child | Total |
|------------|------------------------------|---------------------------|----------|
| Married | O: 176, E:$\frac{324 \cdot 310}{600} = 167.4$| O: 148, E:$\frac{324 \cdot 290}{600} = 156.6$ | 324 |
| Did Not Marry | O: 134, E:$\frac{276 \cdot 310}{600} = 142.6$ | O: 142, E:$\frac{276 \cdot 290}{600} = 133.4$ | 276 |
| Total | 310 | 290 | 600 |

## Calculating the $\chi^2$ test statistic

$\chi^2 = \sum \frac{(Observed - Expected)^2}{Expected}$

$\chi^2 = \frac{176-167.4}{167.4} + \frac{148-156.6}{156.6} + \frac{134-142.6}{142.6} + \frac{143-133.4}{133.4}$

$\chi^2 = 0.441816 + 0.5186536 + 0.4722861 + 0.554428 = 1.987178$

If we compare this to an F distribution with 1 df (http://gallery.shinyapps.io/dist_calc), we see that the p-value is approximately 0.158.  We fail to reject the null hypothesis that the a single mother marrying is independent from her child's gender.

\newpage
## Testing against R Output
```{r cars}
M <- as.table(rbind(c(176,148), c(134,142)))
dimnames(M) <- list(Marital = c("Married","Did Not Marry"),Child = c("Boy", "Girl"))
M
chisq<-chisq.test(M, correct = FALSE)
chisq
```

