---
title: "Likelihoods"
author: "SDS 291 - Multiple Regression"
date: "4/22/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(Stat2Data)
data("Titanic")
```

# Likelihoods

$$L = \prod \hat\pi^{y_i} \times (1-\hat\pi)^{1-y_i}$$

# Titanic Data

## Table of Sex and Survival
```{r}
gmodels::CrossTable(Titanic$Sex, Titanic$Survived, prop.chisq = FALSE, prop.c = FALSE, prop.t = FALSE)
```

### Null  Deviance

What if *everyone* had the same probability of survival, regardless of their sex? Calculate the $\pi$ for the total sample, irrespective of sex and calculate the Log-Likelihood there.

```{r}
TitNullL = (0.343^450) * (0.657^863) 
TitNulllog_L = 450*(log(0.343)) + 863*(log(0.657))
TitNullNeg2LL = -2*(TitNulllog_L)
cbind(`-2LL` = TitNullNeg2LL, LogLikelihood = TitNulllog_L, Likelihood = TitNullL)
```

### Residual Deviance

The alternative is that there *is* some relationship between probability of survival and sex. 

```{r}
TitL = (0.167^142) * (0.833^709) * (.667^308) * (.333^154)
Titlog_L = 142*(log(0.167)) + 709*(log(.833)) + 308*(log(.667)) + 154*(log(0.333)) 
TitNeg2LL = -2*(Titlog_L)
cbind(`-2LL` = TitNeg2LL, LogLikelihood = Titlog_L, Likelihood = TitL)
```

### Regression Models

Do you get the same -2LogLikelihood as the Null and Residual Deviance statistics in the regression model below?  

```{r}
m_titanic<-glm(Survived~SexCode, data=Titanic, family=binomial)
summary(m_titanic)
```

#### *G* Statistic or Drop-in-Deviance test
```{r, error=FALSE, warning=FALSE,message=FALSE}
library(lmtest)
lrtest(m_titanic)
```

Can you identify/calculate the *G* test statistic from the information given in the regression summary?

You can use the code above to test whether *G* is statistically signficant.  Or the anova function:

```{r}
anova(m_titanic, test="Chisq")
```

Alternatively, you can use the general framework of `R` code for `1-pchisq(G,df)` to calculate the p-value where *G* is the *G* test statistic and *df* is the degrees of freedom in the difference between the null and alternative models (in this case, think about how many $\hat\beta$ there are that are different between the null and alternative models -- the number is the degree of freedom).

```{r}
G <- TitNullNeg2LL - TitNeg2LL
1-pchisq(G,1)
```

# `Whickham` Data - Smoking and Survival

```{r, error=FALSE, message=FALSE, warning=FALSE, echo=FALSE}
library(mosaic)
data("Whickham")
gmodels::CrossTable(Whickham$smoker, Whickham$outcome, prop.chisq = FALSE, prop.c = FALSE, prop.t = FALSE)
```

## Do it by Hand

Practice with the example above to calculate the null and residual deviance from this 2-by-2 table.

<!---Null Deviance--->
```{r, echo=FALSE, include=FALSE, eval=FALSE}
table(Whickham$outcome)
```

```{r, echo=FALSE, include=FALSE, eval=FALSE}
pi = 945/(945+369)
pi
```

```{r, echo=FALSE, include=FALSE, eval=FALSE}
NullL = (pi^945) * ((1-pi)^369)
Nulllog_L = 945*(log(0.72)) + 369*(log(.28))
NullNeg2LL = -2*(Nulllog_L)
cbind(`-2LL` = NullNeg2LL, LogLikelihood = Nulllog_L, Likelihood = NullL)
```


<!---Residual Deviance--->
```{r, include=FALSE, echo=FALSE, eval=FALSE}
L = (0.761^443) * (0.239^139) * (0.686^502) * (0.314^230)
log_L = 443*(log(0.761)) + 139*(log(0.239)) + 502*(log(0.686)) + 230*(log(0.314)) 
Neg2LL = -2*(log_L)
cbind(`-2LL` = Neg2LL, LogLikelihood = log_L, Likelihood = L)
```


## Regression Modeling

Can you get the same example as the regression model?
```{r, echo=FALSE, eval=FALSE, include=FALSE}
Whickham$outcome<-relevel(Whickham$outcome, ref="Dead")
m0_Whickham<-glm(outcome~smoker, data=Whickham, family=binomial)
summary(m0_Whickham)
```

## Calculate and Test *G*

Calculate *G* by hand and evaluate that value on the $\chi^2$ distribution with 1 df: (https://gallery.shinyapps.io/dist_calc/)

Use one of the functions above to have R calculate it for you.

```{r, eval=FALSE, echo=FALSE, include=FALSE}
anova(m0_Whickham, test="Chisq")
lrtest(m0_Whickham)
```

Try both!

<!-- ## Visualizing the difference that we're testing with the Likelihoods -->

<!-- Here's a way to generate a figure of this relationship on the probability scale. -->
```{r, eval=FALSE, include=FALSE, echo=FALSE}
library(mosaic)
data("Whickham")
#Here the explanatory variable is smoker and the response variable is outcome (dead/alive)
binary_by_binary <- Whickham %>% #saving the plot as binary_by_binary, working from the Whickham data set 
  group_by(smoker) %>% #by smoking status
  count(outcome) %>% #count how many are dead and alive
  mutate(pi=n/sum(n)) %>% #calculate the count as a proportion (proprtion dead of smokers; dead of nonsmokers)
  filter(outcome=="Dead") %>% #only keep the dead outcomes (alive is just 1-pi)
  ggplot(aes(y=pi, x=smoker)) + #starting the plot of the probability by smoking status
  ylab("Probabiltiy of Dead") + #Just naming the y-xis so it's clear which outcome we're plotting (alive or dead)
  geom_bar(stat="identity") #plot the proportion/probability for each smoker group

binary_by_binary
```