---
title: "Multiple Logistic Regression"
author: "SDS 291"
date: "April 15, 2020"
output:
  pdf_document: default
  word_document: default
---

```{r, warning=FALSE, error=FALSE, message=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We're going to work with the `Whickham` data contains observations about women, and whether they were alive 20 years after their initial observation (`outcome` is a 2 level factor variable - Alive/Dead). You can learn more about these data from the `mosaicData` help feature if you'd like.

Specifically, we're interested in: the association of age, smoking status (smoker), and 20-year survival (outcome: alive (success), dead (reference / failure)).  Bring in the relevant packages and the data (below, from the mosaic package).

```{r, warning=FALSE, error=FALSE, message=FALSE}
library(mosaic)
library(tidyverse)
data("Whickham")
Whickham$outcome<-relevel(Whickham$outcome, ref="Dead")
```

\newpage

# Age and Outcome

```{r, fig.height=2, fig.width=3, echo=FALSE}
qplot(x=outcome,y=age,data=Whickham)+geom_jitter()+coord_flip()
qplot(x=outcome,y=age,data=Whickham,geom="boxplot")+coord_flip()
```


```{r}
m0<-glm(outcome~age, data=Whickham, family=binomial)
summary(m0)
```

1. Write the fitted equation and interpret the results (in these units) in light of the question. Be sure to comment on the magnitude and direction of the association.

\vspace{1in}

2. Based on this model, what is the probability that a 60 year old was alive 20 years after the initial survey?

\vspace{2in}

\newpage
# Smoking Status and Outcome (Alive)

```{r}
Whickham$smoker<-factor(Whickham$smoker, levels=c("Yes", "No"))
Whickham$outcome<-factor(Whickham$outcome, levels=c("Alive", "Dead"))
tally(~ smoker + outcome, margins=FALSE, data=Whickham)
```


3. Calculate the Odds Ratio of smokers being alive in 20 years compared to non-smokers from the table above.
\vspace{3in}

```{r}
Whickham$smoker<-relevel(Whickham$smoker, ref= "No")
Whickham$outcome<-relevel(Whickham$outcome, ref= "Dead")
m1<-glm(outcome~smoker, data=Whickham, family=binomial)
summary(m1)
```

4. Show that you can calculate the coefficient for smoking status from your regression model as you did in #3.

\vspace{3in}

5. Based on your model, what's the probability that a smoker was alive 20 years later?

\vspace{3in}

6. Based on what you know about the risk of death for age and smoking status, do these results make sense? Explain your answer.

\newpage
## Multiple Logistic Regression
```{r}
m2<-glm(outcome~age+smoker, data=Whickham, family=binomial)
summary(m2)
```

7. What is the odds ratio for smokers compared to non-smokers in this model? Interpret in a sentence in the context of this real-world problem.

\vspace{3in}


8. What is the probability of a 60 year old non-smoker being alive 20 years later?

\vspace{3in}


9. What is the probability of a 40 year old smoker being alive 20 years later?

\vspace{3in}


10. What does this model help us to understand about our simple logistic regression estimates above?

\vspace{3in}


### _Optional_ - Interaction Term

11. What would an interaction term between age and smoking status do in this model? How would an interaction term affect the OR for age?

\vspace{3in}

12. How do the coefficients in the interaction model relate to the separate models for Age for smokers and non-smokers (below)?

\vspace{0.5in}

#### Interaction 
```{r}
m3<-glm(outcome~age+smoker+age*smoker, data=Whickham, family=binomial)
summary(m3)
```



#### Smokers
```{r}
Whickham_smoker<- Whickham %>% filter(smoker=="Yes")
m3_smoker<-glm(outcome~age, data=Whickham_smoker, family=binomial)
summary(m3_smoker)
```

#### Non-Smokers
```{r}
Whickham_nonsmoker<- Whickham %>% filter(smoker=="No")
m3_nonsmoker<-glm(outcome~age, data=Whickham_nonsmoker, family=binomial)
summary(m3_nonsmoker)
```

\vspace{3in}

13. Is this model with the interaction term a better fit than the model with Age alone (Model 0 above), or than the model with just Smoking alone (Model 1)?

```{r, error=FALSE, message=FALSE, warning=FALSE}
library(lmtest)
lrtest(m3,m0)
lrtest(m3,m1)
```

a. What are the null and alternative hypotheses for each of these tests?
\vspace{2in}

b. What is the test statistic and p-value for each and what does that mean about the test?

\vspace{2in}

c. What do these tests tell you about the relationships between age, smoking, and survival over 20 years in this cohort of women from Whickham?
\vspace{2in}