---
title: "SDS 291 - Multiple Regression - Day 02"
date: "January 29, 2020"
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(Stat2Data)
data("LeafWidth")
```

Biologists know that the leaves on plants tend to get smaller as temperatires rise. The dataset `LeafWidth` has fata on samples of leaves from the species _Dodonaea viscosa_ subsp, _angustissima_, which have been collected in a certain region of South Australia for man years.

The variable `Width` is the average width, in mm, of leaves, taken at their widest points, that wer collected in a given year.


## 1. Scatterplot of Leaf Width (mm) and Year

```{r scatter, echo=FALSE, fig.height=4}
qplot(x=Year, y=Width, data=LeafWidth)
```

### Describe the scatterplot in words.




## 2. Find the least squares regression line for predicting leaf width based on year

```{r reg, echo=FALSE, include=FALSE}
model_width<-lm(Width~Year, data=LeafWidth)
summary(model_width)
```


### Write the fitted regression model

### Interpret the value of the slope for the fitted model in the context of this setting.




## 3. Assessing the model

## What leaf width would the fitted model predict a leaf in 1994 would have?

## Find the residual 

[hint: do the calculation of the fitted value by hand, find the actual number from the data, and calculate the residual from there]

```{r residual, echo=FALSE}
LeafWidth %>%
  filter(Year==1994) %>%
  head()
```


## 4. Visualize the model


### Make a scatterplot that includes the regression line

```{r scatter, echo=FALSE, fig.height=4}
qplot(x=Year, y=Width, data=LeafWidth) + geom_smooth(method="lm", se=FALSE)
```


## Make a histogram of the residuals

```{r scatter, echo=FALSE, fig.height=4, message=FALSE}
qplot(x=model_width$residuals, geom="histogram")
```

## Make a probability plot and residual fitted plot

```{r scatter, echo=FALSE, fig.height=4, message=FALSE}
plot(model_width, which=c(1:2))
```
