---
title: "MLR Examples"
author: "SDS 291"
date: "2/24/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(tidyverse)
library(Stat2Data)
data("RailsTrails")
```

# Rail Trail Multiple Regression Example
We're still using data from a sample of 104 homes in Northampton, MA to see whether being close to the bike trail enhances the value of the home. Specifically, we're looking at the association between square feet (a house's size) and distance from the rail trail with the house's estimated value in 2014. The variables we're using are: 

- `Price2014`: Zillow price estimate from 2014 (in thousands of dollars)
- `Distance`: Distance (in miles) to the nearest entry point to the rail trail network
- `SquareFeet`: Square footage of interior finished space (in thousands of sf)

```{r regression line}
library(Stat2Data)
data("RailsTrails")
m1<-lm(Price2014 ~ SquareFeet + Distance , data = RailsTrails)
summary(m1)
```

$\widehat{Price2014} = \hat{\beta}_0 + \hat{\beta}_1 \cdot SquareFeet + \hat{\beta}_2 \cdot Distance$

1. What price would this model predict for a 1000 square foot house that is _1 mile_ from the rail trail? (Be cautious with the units)

```{r, echo=FALSE}
PriceM1_1mi<-78.985 + 147.920 -15.788
```

$\widehat{Price2014} = 78.985 + 147.920 \cdot 1 -15.788 \cdot 1=$ `r PriceM1_1mi`


2. What price would this model predict for a 1000 square foot house that is _2 miles_ from the rail trail? (Be cautious with the units)

```{r, echo=FALSE}
PriceM1_2mi<-78.985 + 147.920 -(15.788*2)
```

$\widehat{Price2014} = 78.985 + 147.920 \cdot 1 -15.788 \cdot 2=$ `r PriceM1_2mi`

```{r, echo=FALSE}
#install.packages("scatterplot3d") # Install
library("scatterplot3d") # load
RailsTrails_2<-RailsTrails %>%
  select(SquareFeet,Distance,Price2014)
# 3D scatter plot
s3d <- scatterplot3d(RailsTrails_2, type = "h", color = "blue", angle=20, pch = 16)
# Add regression plane
my.lm <- lm(RailsTrails_2$Price2014 ~ RailsTrails_2$SquareFeet + RailsTrails_2$Distance)
s3d$plane3d(my.lm)
```
\vspace{0.5 in}

|     | 1 mile  | 2 miles | Difference |
|-----|---------|---------| ---------|
| 1000 $ft^2$ | `r PriceM1_1mi` | `r PriceM1_2mi` | `r PriceM1_2mi-PriceM1_1mi`

For every 1 additional mile away from the Rail Trail, the price of a house in 2014 would decrease by $15,788, on average, adjusted for house size in square feet.

\newpage

# Adjusting for Distance Group

Rather than distance in miles, what if we thought a more useful measure would be whether the house was closer (<1 mile from an entrance to the rail trail) or further away ($\geq$ 1 mile from a rail trail entrance)?

- `Price2014`: Zillow price estimate from 2014 (in thousands of dollars)
- `DistGroup`: 
    - `Closer`: <1 mile to the nearest entry point to the rail trail network
    - `Farther Away`: $\geq$ 1 mile to the nearest entry point to the rail trail network
- `SquareFeet`: Square footage of interior finished space (in thousands of sf)

`R` treats `DistGroup` as a `factor` variable. It can also be treated as a numeric variable, where one category has the value of 0 and the other category has the value of 1. In other words, you can think that the numerical equivalent is: "Closer" = 0 and "Farther Away" = 1.

```{r Distance Group}
m2<-lm(Price2014 ~ SquareFeet+DistGroup , data = RailsTrails)
summary(m2)
```

**$$\widehat{Price2014} = \hat{\beta}_0 + \hat{\beta}_1 \cdot SquareFeet + \hat{\beta}_2 \cdot Far$$**

1. What price would this model predict for a 1000 square foot house that is _Closer_ from the rail trail? 

```{r, echo=FALSE}
PriceM2_1000closer<-80.10 + 150.50
```

$\widehat{Price2014} = 80.10 + 150.50 \cdot 1 -36.97 \cdot 0 =$ `r PriceM2_1000closer`


2. What price would this model predict for a 1000 square foot house that is _Farther Away_ from the rail trail?

```{r, echo=FALSE}
PriceM2_1000far<-80.10 + 150.50 -36.97
```

$\widehat{Price2014} = 80.10 + 150.50 \cdot 1 -36.97 \cdot 1=$ `r PriceM2_1000far`


3. What price would this model predict for a 2000 square foot house that is _Closer_ from the rail trail? 

```{r, echo=FALSE}
PriceM2_2000closer<-80.10 + (150.50*2)
```

$\widehat{Price2014} = 80.10 + 150.50 \cdot 2 -36.97 \cdot 0=$ `r PriceM2_2000closer`


4. What price would this model predict for a 2000 square foot house that is _Farther Away_ from the rail trail?

```{r, echo=FALSE}
PriceM2_2000far<-80.10 + (150.50*2) -36.97
```

$\widehat{Price2014} = 80.10 + 150.50 \cdot 2 -36.97 \cdot 1=$ `r PriceM2_2000far`

```{r, echo=TRUE, warning=FALSE, message=FALSE, fig.height=4}
library(moderndive)
qplot(y=Price2014, x=SquareFeet, color=DistGroup, data=RailsTrails)+geom_parallel_slopes(se=FALSE)
```

\vspace{0.5 in}

| $ft^2$  | Close  | Far | Difference |
|----------|---------|---------| ---------|
| 1000 $ft^2$ | `r PriceM2_1000closer` | `r PriceM2_1000far` | `r PriceM2_1000far-PriceM2_1000closer` |
| 2000 $ft^2$ | `r PriceM2_2000closer` | `r PriceM2_2000far` | `r PriceM2_2000far-PriceM2_2000closer` |
| Difference | `r PriceM2_2000closer-PriceM2_1000closer` | `r PriceM2_2000far-PriceM2_1000far` |  |

**For every additional 1000 $ft^2$ of house size, our model suggests the estimated 2014 price will increase, on average, by $`r prettyNum(((PriceM2_2000closer-PriceM2_1000closer)*1000), big.mark = ",")`, adjusted for whether the house is closer or farther from the rail trail.**

**On average, a house that is farther away from the rail trail will be $`r prettyNum(((PriceM2_1000far-PriceM2_1000closer)*1000), big.mark = ",")` less than a house closer to the rail trail, adjusted for house size in square feet.**

The slope of the line is defined by the house size ($ft^2$); the distance (far vs. close) affects the intercept. So the lines have parallel slopes, and the difference between two identically sized houses where one is close and one that is far will always be the same amount regardless of how big or small the house is.


# Bedrooms

Rather than square feet, let's consider the number of bedrooms the house has, in addition to its distance from the rail trail.

- `Price2014`: Zillow price estimate from 2014 (in thousands of dollars)
- `BedGroup`: Categorical Variable of house type by group of bedrooms:
    - 1-2 bedrooms (reference), 
    - 3 bedrooms, 
    - 4+ bedrooms
- `Distance`: Distance (in miles) to the nearest entry point to the rail trail network

You can think about `BedGroup` similarly to `DistGroup` and consider the 3 bedroom group output in the model below akin to an indicator variable with the values of 0 or 1: 0 if the house doesn't have 3 bedrooms and 1 if it does have 3 bedrooms. Same for 4+ bedrooms.

```{r}
m3 <- lm(Price2014 ~ Distance+BedGroup, data = RailsTrails)
summary(m3)
```

$\widehat{Price2014} = \hat{\beta}_0 + \hat{\beta}_1 \cdot Distance + \hat{\beta}_2 \cdot 3 Beds + \hat{\beta}_3 \cdot {4+} Beds$

## 1 mile Distance
5. What price would this model predict for a house 1 mile away that has 1-2 Bedrooms? 

```{r, echo=FALSE}
PriceM3_1mile1_2beds<-283.56 -42.30
```

$\widehat{Price2014} = 283.56 - 42.30 \cdot 1 + 39.88 \cdot 0 + 106.13 \cdot 0=$ `r PriceM3_1mile1_2beds`


6. What price would this model predict for a house 1 mile away that has 3 Bedrooms?

```{r, echo=FALSE}
PriceM3_1mile3beds<-283.56 -42.30 + 39.88
```

$\widehat{Price2014} = 283.56 - 42.30 \cdot 1 + 39.88 \cdot 1 + 106.13 \cdot 0=$ `r PriceM3_1mile3beds`

7. What price would this model predict for a house 1 mile away that has 4+ Bedrooms?

```{r, echo=FALSE}
PriceM3_1mile4beds<-283.56 - 42.30 + 106.13
```

$\widehat{Price2014} = 283.56 - 42.30 \cdot 1 + 39.88 \cdot 0 + 106.13 \cdot 1=$ `r PriceM3_1mile4beds`

## 2 mile Distance

8. What price would this model predict for a house 2 miles away that has 1-2 Bedrooms? 

```{r, echo=FALSE}
PriceM3_2mile1_2beds<-283.56 +(-42.30*2)
```

$\widehat{Price2014} = 283.56 - 42.30 \cdot 2 + 39.88 \cdot 0 + 106.13 \cdot 0=$ `r PriceM3_2mile1_2beds`


9. What price would this model predict for a house 2 miles away that has 3 Bedrooms?

```{r, echo=FALSE}
PriceM3_2mile3beds<-283.56  +(-42.30*2) + 39.88
```

$\widehat{Price2014} =283.56 - 42.30 \cdot 2 + 39.88 \cdot 1 + 106.13 \cdot 0 =$ `r PriceM3_2mile3beds`

10. What price would this model predict for a house 2 miles away that has 4+ Bedrooms?
```{r, echo=FALSE}
PriceM3_2mile4beds<-283.56 + (-42.30*2) + 106.13
```

$\widehat{Price2014} = 283.56 - 42.30 \cdot 2 + 39.88 \cdot 0 + 106.13 \cdot 1 =$ `r PriceM3_2mile4beds`


```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.height=3.5}
#calculating the different intercepts
intercepts <- c(coef(m3)["(Intercept)"],
                coef(m3)["(Intercept)"] + coef(m3)["BedGroup3 beds"],
                coef(m3)["(Intercept)"] + coef(m3)["BedGroup4+ beds"])

#putting these into a new data frame with the slope and transmission indicator variable
lines.df <- data.frame(intercepts = intercepts,
                       slopes = rep(coef(m3)["Distance"], 3),
                       bedgroup = as.factor(levels(RailsTrails$BedGroup)))
#making the plot
qplot(x = Distance, y = Price2014, color = BedGroup, data = RailsTrails) + 
  geom_abline(aes(intercept = intercepts, 
                  slope = slopes, 
                  color = bedgroup), data = lines.df)
```





| Distance  | 1-2 Beds  | 3 beds  | Difference |
|----------|---------|---------| ---------|
| 1 mile | `r PriceM3_1mile1_2beds` |  `r PriceM3_1mile3beds` | `r PriceM3_1mile3beds-PriceM3_1mile1_2beds` |
| 2 miles | `r PriceM3_2mile1_2beds` |  `r PriceM3_2mile3beds` | `r PriceM3_2mile3beds-PriceM3_2mile1_2beds` |
| Difference | `r PriceM3_2mile1_2beds-PriceM3_1mile1_2beds` | `r PriceM3_2mile3beds-PriceM3_1mile3beds` |  |

| Distance  | 1-2 Beds  | 4+ beds  | Difference |
|----------|---------|---------| ---------|
| 1 mile | `r PriceM3_1mile1_2beds` |  `r PriceM3_1mile4beds` | `r PriceM3_1mile4beds-PriceM3_1mile1_2beds` |
| 2 miles | `r PriceM3_2mile1_2beds` |  `r PriceM3_2mile4beds` | `r PriceM3_2mile4beds-PriceM3_2mile1_2beds` |
| Difference | `r PriceM3_2mile1_2beds-PriceM3_1mile1_2beds` | `r PriceM3_2mile4beds-PriceM3_1mile4beds` |  |

**Similar to the example above with a binary explanatory variable, the bed group categorical variable also affects the intercepts, which makes parallel lines for each of the three bedroom groups with the same slope based on distance to the rail trail.**

\newpage

# Back to Distance Group

What if we thought that the relationship between Square Footage of a house and its price would *vary* by whether it's closer or further from the rail trail.  A big house may not matter as much if it's really far from the rail trail, and a smaller house may be more valuable if it's closer to the rail trail than if it were further away. 

You can also think about bigger houses in terms of an addition to a house -- will adding another 500$ft^2$ increase the value of your price the same amount regardless of where the house is located, or do you think that adding 500$ft^2$ to a house close to the rail trail will matter more/less for the total house's price than if the house were further from the rail trail.  Imagine you were a realtor and were asked whether a homeowner should build an addition to their house -- how would you respond?

```{r}
m4 <- lm(Price2014 ~ SquareFeet*DistGroup , data = RailsTrails)
summary(m4)
```

**$$\widehat{Price2014} = \hat{\beta}_0 + \hat{\beta}_1 \cdot SquareFeet + \hat{\beta}_2 \cdot FartherAway + \hat{\beta}_3 \cdot (SquareFeet \times FartherAway)$$**

11. What price would this model predict for a 1000 square foot house that is _Closer_ from the rail trail? 

```{r, echo=FALSE}
PriceM4_1000Closer<-32.16 + 177.82
```

$\widehat{Price2014} = 32.16 + 177.82 \cdot 1 + 32.46 \cdot 0 - 42.15 \cdot 0=$ `r PriceM4_1000Closer`

12. What price would this model predict for a 1000 square foot house that is _Farther Away_ from the rail trail?


```{r, echo=FALSE}
PriceM4_1000Far<-32.16 + 177.82 + 32.46 - 42.15
```

$\widehat{Price2014} = 32.16 + 177.82 \cdot 1 + 32.46 \cdot 1 - 42.15 \cdot 1=$ `r PriceM4_1000Far`

13. What price would this model predict for a 2000 square foot house that is _Closer_ from the rail trail? 

```{r, echo=FALSE}
PriceM4_2000Closer<-32.16 + (177.82*2)
```

$\widehat{Price2014} = 32.16 + 177.82 \cdot 2 + 32.46 \cdot 0 - 42.15 \cdot 0=$ `r PriceM4_2000Closer`

14. What price would this model predict for a 2000 square foot house that is _Farther Away_ from the rail trail?

```{r, echo=FALSE}
PriceM4_2000Far<-32.16 + (177.82*2) + 32.46 - (42.15*2)
```

$\widehat{Price2014} = 32.16 + 177.82 \cdot 1 + 32.46 \cdot 1 - 42.15 \cdot 2=$ `r PriceM4_2000Far`

\newpage

```{r}
qplot(y=Price2014, x=SquareFeet, data=RailsTrails, color=DistGroup) + 
  geom_smooth(method=lm,se=FALSE,fullrange = TRUE)
```

\vspace{0.5 in}

| $ft^2$  | Close  | Far | Difference |
|----------|---------|---------| ---------|
| 1000 $ft^2$ | `r PriceM4_1000Closer` | `r PriceM4_1000Far` | `r PriceM4_1000Far-PriceM4_1000Closer` |
| 2000 $ft^2$ | `r PriceM4_2000Closer` | `r PriceM4_2000Far` | `r PriceM4_2000Far-PriceM4_2000Closer` |
| Difference | `r PriceM4_2000Closer-PriceM4_1000Closer` | `r PriceM4_2000Far-PriceM4_1000Far` |  |


**The model suggests that the relationship between SquareFeet and Price varies by whether the house is close or far from the rail trail. (This difference is not statistically significant, but let's not worry about that for now; let's just think about the coefficients and what they are telling us separate from their hypothesis tests.)**

The $\hat{\beta}_2 Far$ variable affects the intercept. A house that has 0 square feet and is far from the rail trail will be worth $`r prettyNum(coef(m4)["DistGroupFarther Away"]*1000, big.mark = ",")` more than a house of the same size that is closer.

For every additional 1000 $ft^2$ of house size, our model suggests the estimated 2014 price will increase, on average, though at different rates depending on whether the house is close or far from the rail trail. Houses far from the rail trail will increase more slowly in price as they have more square footage than houses closer to the rail trail do.

For every additional 1000 $ft^2$ in size, the price of a house that is _close_ to the rail trail will, on average, increase by $`r prettyNum(((coef(m4)["SquareFeet"])*1000), big.mark = ",")`.

For every additional 1000 $ft^2$ in size, the price of a house that is _far_ to the rail trail will, on average, increase by $`r prettyNum(((coef(m4)["SquareFeet"]+coef(m4)["SquareFeet:DistGroupFarther Away"])*1000), big.mark = ",")`.

The _marginal difference_ in the house price for each 1000 square feet between houses that are far versus those that are closer is, on average,  $`r prettyNum(((coef(m4)["SquareFeet:DistGroupFarther Away"])*1000), big.mark = ",")`
