---
title: "Logistic Regression"
author: "SDS 291"
date: "April 13, 2020"
output:
  pdf_document: default
  word_document: default
---

```{r, warning=FALSE, error=FALSE, message=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(mosaic)
library(Stat2Data)
library(magrittr)
library(tidyverse)
data("Whickham")
Whickham$outcome<-relevel(Whickham$outcome, ref="Dead")
```

We're going to work with the `Whickham` data contains observations about women, and whether they were alive 20 years after their initial observation (`outcome` is a 2 level factor variable - Alive/Dead). You can learn more about these data from the `mosaicData` help feature if you'd like.

Specifically, we're interested in: the association of age, smoking status (smoker), and 20-year survival (outcome: alive (success), dead (reference / failure)).  Bring in the relevant packages and the data (above, from the mosaic package).

# Age and Outcome

```{r, fig.height=3, fig.width=3, echo=FALSE}
qplot(x=outcome,y=age,data=Whickham)+geom_jitter()+coord_flip()
qplot(x=outcome,y=age,data=Whickham,geom="boxplot")+coord_flip()
```


```{r}
m0<-glm(outcome~age, data=Whickham, family=binomial)
summary(m0)
```

1. Write the fitted equation and interpret the results (in these units) in light of the question. Be sure to comment on the magnitude and direction of the association.

\vspace{3in}

2. Based on this model, what is the probability that a 60 year old was alive 20 years after the initial survey?

\vspace{3in}

\newpage
# Smoking Status and Outcome (Alive)

```{r}
Whickham$smoker<-factor(Whickham$smoker, levels=c("Yes", "No"))
Whickham$outcome<-factor(Whickham$outcome, levels=c("Alive", "Dead"))
tally(~ smoker + outcome, margins=FALSE, data=Whickham)
```


3. Calculate the Odds Ratio of non-smokers being alive in 20 years compared to smokers from the table above.
\vspace{3in}

```{r}
Whickham$smoker<-relevel(Whickham$smoker, ref= "No")
Whickham$outcome<-relevel(Whickham$outcome, ref= "Dead")
m1<-glm(outcome~smoker, data=Whickham, family=binomial)
summary(m1)
```

4. Show that you can calculate the coefficient for smoking status from your regression model as you did in #3.

\vspace{3in}

5. Based on your model, what's the probability that a smoker was alive 20 years later?

\vspace{3in}

6. Based on what you know about the risk of death for age and smoking status, do these results make sense? Explain your answer.

\newpage
## Multiple Logistic Regression
```{r}
m2<-glm(outcome~age+smoker, data=Whickham, family=binomial)
summary(m2)
```

7. What is the odds ratio for smokers compared to non-smokers in this model? Interpret in a sentence in the context of this real-world problem.

\vspace{3in}


8. What is the probability of a 60 year old non-smoker being alive 20 years later?

\vspace{3in}


9. What is the probability of a 40 year old smoker being alive 20 years later?

\vspace{3in}


10. What does this model help us to understand about our simple logistic regression estimates above?

\vspace{3in}
