---
title: "Simple Logistic Regression Lab"
author: "SDS 291"
date: "April 8, 2020"
output:
  html_document:
    css: "~/Dropbox/3_Smith/SDS291/sds291_site/static/img/labs.css"
    highlight: textmate
    theme: journal
    toc: true
---

```{r, eval=FALSE, warning=FALSE, error=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(mosaic)
library(Stat2Data)
library(magrittr)
library(tidyverse)
```

We're going to keep working with the `Whickham` data from class on Tuesday.  Bring in the relevant packages (above) and the data (below, from the mosaic package).

```{r, eval=FALSE}
data("Whickham")
?Whickham
```

# Age and Outcome

1. Visualize the relationship between age and the outcome

```{r, eval=FALSE}
qplot(x=age,y=outcome,data=Whickham, geom="boxplot")
qplot(x=outcome,y=age,data=Whickham)+geom_jitter()+coord_flip()
qplot(x=outcome,y=age,data=Whickham,geom="boxplot")+coord_flip()
```

1. Fit a logistic model to test the relationship between outcome and age (with code below).



```{r, eval=FALSE}
Whickham$outcome<-relevel(Whickham$outcome, ref="Alive")
m0<-glm(outcome~age, data=Whickham, family=binomial)
summary(m0)
```

`relevel` establishes the reference group (`ref=`) for a given variable. Remember, `R`'s default is to use the first alphabetical as the reference group, which would mean non-smokers and dead would be the reference groups. Instead, we want to flip that around.  This command uses "base" `R`, which uses the `dataset$variablename` syntax.

1. Write the fitted equation and interpret the results (in these units) in light of the question. Be sure to comment on the magnitude and direction of the association.

1. Based on this model, what is the probability that a 60 year old was dead?

```{r, eval=FALSE}
newdata = data.frame(age=60)
predict(m0, newdata, type="response")

age<-60
logodds<-coef(m0)[1]+ (coef(m0)[2]*age)
m0_data<-as.data.frame(cbind(age,logodds))
m0_data <- m0_data %>%
  mutate(odds=exp(logodds),
        prob = odds/(1+odds))
m0_data
```

1. What is the odds ratio for the association between age and dying?

```{r, eval=FALSE}
exp(coef(m0))
```

# Outcome and Smoker

## 2-by-2 table
Construct a two-way table to explore whether smoking is associated with the outcome (staying alive).

```{r, eval=FALSE}
Whickham$smoker<-relevel(Whickham$smoker, "Yes", "No")
Whickham$outcome<-relevel(Whickham$outcome, "Dead","Alive")
tally(~ smoker + outcome, margins=FALSE, data=Whickham)
tally(~ smoker + outcome, margins=TRUE, format = "proportion", data=Whickham)
```


1. Calculate the the proportion of smokers who were dead and of smokers who were alive 20 years later. 


1. Use the code below to calculate the Odds Ratio using the numbers of people in the relevant table cells (a-d).

```{r, eval=FALSE}
#         Response
#        Yes | No
# Expl.|-----|-----|
# Yes  |  a  |  b  |
#      |-----|-----|
# No   |  c  |  d  |
#      |-----|-----|
# 
OR<-(a/b)/(c/d)
OR
```


## Modeling

First we have to be sure that we are going to calculate the same OR that we did above -- smokers' odds of dying compared to non-smokers'.

1. Relevel the `smoker` variable with the appropriate reference group


1. Fit a logistic model of the relationship between smoking status and outcome.
```{r, eval=FALSE}
Whickham$outcome<-relevel(Whickham$outcome, ref="Alive")
Whickham$smoker<-relevel(Whickham$smoker, ref="No")
m1<-glm(outcome~smoker, data=Whickham, family=binomial)
summary(m1)
exp(cbind(OR=coef(m1), confint(m1)))
```

1. Write the fitted equation and interpret the results (in these units) in light of the question. Be sure to comment on the magnitude and direction of the association.

1. Using the two-way table above, show that you can calculate the coefficient for smoking status from your regression model.

1. Based on your model, what's the probability that a smoker was alive?

1. What are the regression assumptions? Evaluate whether they are met.

[Hint: for the question above, check the textbook in Chapter 9.3-9.4 for what the assumptions are.  See the code below for one of the assumptions (linearity, p.473). The book suggest ~5 bins, but we're using 10]
```{r, eval=FALSE, error=FALSE, warning=FALSE, message=FALSE}
#let's make `outcome` a numeric variable between 0 and 1 so that the proportion makes sense
#and decide what the 10 age groups should be by "cutting" the data into 10 evenly sized groups.
Whickham <- Whickham %>%
  mutate(isDead=as.numeric(if_else(outcome=="Dead",1,0)),
         ageGroup = cut(age, breaks=10))

# Then generate the proportion of those alive for each age group (so you have something to plot on the y-axis)
# And the average age for each age group (so you have something to plot on the x-axis)
Whickham_bin <- Whickham %>%
  group_by(ageGroup) %>%
  mutate(binned.y=mean(isDead), binned.x=mean(age))

#is this roughly linear on the log-odds scale?
qplot(y=logit(binned.y), x=binned.x, data=Whickham_bin) + geom_smooth(method="lm", se=FALSE, fullrange=TRUE)

```