---
title: "Model Inference and Partitioning Error"
author: "Benjamin D. Capistrant, Smith College"
date: "Feb. 5, 2020"
output:
  sds::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r xaringan-themer, include=FALSE}
# sds::duo_smith()
sds::mono_light_smith()
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(dplyr)
library(ggplot2)
library(latex2exp)
library(Stat2Data)
data("AccordPrice")
fit <- lm(Price ~ Mileage, data = AccordPrice)
AccordPrice <- AccordPrice %>%
  mutate(predicted = predict(fit),
         residuals = residuals(fit),
         top_right = Mileage + residuals,
         bottom_right = Mileage + residuals,
         y_bar = mean(Price)
    )

```


# Residuals

.pull-left[
```{r, warning=FALSE,error=FALSE,message=FALSE}
# Residual - Overall
ggplot(AccordPrice, aes(x=Mileage, y=Price)) +  # Set up canvas with outcome variable on y-axis
  theme_bw() +
  geom_point() + # Plot the actual points
  geom_point(aes(x=Mileage, y = predicted), color="blue", size=2) + 
  #geom_point(aes(x = top_right, y = Price), shape = 1, color="white") + 
  #geom_point(aes(x = bottom_right, y = predicted), shape = 1, color="white") +
  geom_smooth(method="lm", se=FALSE, size=0.5) +  
  geom_segment(aes(x=Mileage, y=Price, xend = Mileage, yend = predicted)) +
  #coord_cartesian(xlim = c(0,20), ylim=c(15,35)) + 
  annotate("text",x=4.5, y=27, hjust=1, vjust=0.5,
            label=TeX("$y$"), size=8) +
  annotate("text",x=4.5, y=20, hjust=1, vjust=1,
            label=TeX("$\\hat{y}$"), color="blue", size=8)  

```
]

.pull-right[
```{r, warning=FALSE,error=FALSE,message=FALSE}
# Residual - ZOOMED
ggplot(AccordPrice, aes(x=Mileage, y=Price)) +  # Set up canvas with outcome variable on y-axis
  theme_bw() +
  geom_point() + # Plot the actual points
  geom_point(aes(x=Mileage, y = predicted), color="blue", size=2) + 
  geom_point(aes(x = top_right, y = Price), shape = 1, color="white") + 
  geom_point(aes(x = bottom_right, y = predicted), shape = 1, color="white") +
  geom_smooth(method="lm", se=FALSE, size=0.5) +  
  geom_segment(aes(x=Mileage, y=Price, xend = Mileage, yend = predicted)) +
  coord_cartesian(xlim = c(0,20), ylim=c(15,35)) + 
  annotate("text",x=4.5, y=27, hjust=1, vjust=0.5,
            label=TeX("$y$"), size=8) +
  annotate("text",x=4.5, y=20, hjust=1, vjust=1,
            label=TeX("$\\hat{y}$"), color="blue", size=8)   
```

]

---
# Sum of Squared Residuals

```{r, warning=FALSE,error=FALSE,message=FALSE}
# Squared Residuals
ggplot(AccordPrice, aes(x=Mileage, y=Price)) +  # Set up canvas with outcome variable on y-axis
  theme_bw() +
  geom_point() + # Plot the actual points
  geom_point(aes(x=Mileage, y = predicted), color="blue", size=2) + 
  geom_point(aes(x = top_right, y = Price), shape = 1, color="white") + 
  geom_point(aes(x = bottom_right, y = predicted), shape = 1, color="white") +
  geom_smooth(method="lm", se=FALSE, size=0.5) +  
  geom_segment(aes(x=Mileage, y=Price, xend = Mileage, yend = predicted)) +
  coord_cartesian(xlim = c(0,20), ylim=c(15,35)) + 
  annotate("text",x=4.5, y=27, hjust=1, vjust=0.5,
            label=TeX("$y$"), size=8) +
  annotate("text",x=4.5, y=20, hjust=1, vjust=1,
            label=TeX("$\\hat{y}$"), color="blue", size=8) + 
  geom_segment(aes(x=Mileage, y=Price, yend = Price, xend = top_right), color = "grey", linetype = "dashed") +   geom_segment(aes(x=bottom_right, y=predicted, yend = Price, xend = top_right), color = "grey", linetype = "dashed") + 
   geom_segment(aes(x=Mileage, y=predicted, yend = predicted, xend = bottom_right), color = "lightgrey", linetype = "dashed") +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank())
```

---
class: middle, inverse, center

# Separating Model Variance

## SSTotal = SSModel + SSError

## $\sum(y-\bar{y})^2 = \sum(\hat{y}-\bar{y})^2 + \sum(y-\hat{y})^2$
---
# Sum of Squares Total

```{r, error=FALSE, message=FALSE, warning=FALSE}
ggplot(AccordPrice, aes(x=Mileage, y=Price)) +  # Set up canvas with outcome variable on y-axis
  theme_bw() +
  geom_point() + # Plot the actual points
  #geom_point(aes(x=Mileage, y = y_bar), color="red", size=1) + 
  #geom_point(aes(x=Mileage, y = predicted), color="blue", size=1) + 
  geom_smooth(method="lm", se=FALSE, size=0.5) +  
  #geom_hline(yintercept=14.27667, color="red") +
  geom_segment(aes(x=3 ,xend=150,y=AccordPrice$y_bar,yend=AccordPrice$y_bar), color="red") +
  geom_segment(aes(x=Mileage, y=Price, xend = Mileage, yend = AccordPrice$y_bar), color="black", linetype="dashed", size=0.5) +
  annotate("text",x=-.3, y=27, hjust=1, vjust=0.5,
            label=TeX("$y_i$"), size=8) +
  annotate("text",x=-.3, y=20, hjust=1, vjust=0.25,
            label=TeX("$\\hat{y}_i$"), color="blue", size=8) + 
  annotate("text",x=-.3, y=14, hjust=1, vjust=0.5,
            label=TeX("$\\bar{y}$"), color="red", size=8) + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank())
```

---
# Partitioning SST into SSM & SSE

```{r, warning=FALSE, message=FALSE, error=FALSE}

ggplot(AccordPrice, aes(x=Mileage, y=Price)) +  # Set up canvas with outcome variable on y-axis
  theme_bw() +
  geom_point() + # Plot the actual points
  #geom_point(aes(x=Mileage, y = y_bar), color="red", size=1) + 
  #geom_point(aes(x=Mileage, y = predicted), color="blue", size=1) + 
  geom_smooth(method="lm", se=FALSE, size=0.5) +  
  #geom_hline(yintercept=14.27667, color="red") +
  geom_segment(aes(x=3 ,xend=150,y=AccordPrice$y_bar,yend=AccordPrice$y_bar), color="red") +
  geom_segment(aes(x=Mileage, y=Price, xend = Mileage, yend = predicted), color="blue", linetype="solid", size=0.5) +
  geom_segment(aes(x=Mileage, y=predicted, xend = Mileage, yend = AccordPrice$y_bar), color="red", linetype="dashed", size=0.5) +
  annotate("text",x=-.3, y=27, hjust=1, vjust=0.5,
            label=TeX("$y_i$"), size=8) +
  annotate("text",x=-.3, y=20, hjust=1, vjust=0.25,
            label=TeX("$\\hat{y}_i$"), color="blue", size=8) + 
  annotate("text",x=-.3, y=14, hjust=1, vjust=0.5,
            label=TeX("$\\bar{y}$"), color="red", size=8) + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank())

```

---
# If the relationship were weak

```{r, warning=FALSE, message=FALSE, error=FALSE}
library(MASS)
weak <- as.data.frame(mvrnorm(1000, mu = c(0,0), Sigma = matrix(c(1,0.13,0.13,1), ncol = 2),
               empirical = TRUE))
fitweak <- lm(V1 ~ V2, data = weak)
weak <- weak %>%
  mutate(predicted = predict(fitweak),
         residuals = residuals(fitweak),
         y_bar = mean(V1)
    ) %>%
  rename(x=V2, y=V1)


ggplot(weak, aes(x=x, y=y)) +  # Set up canvas with outcome variable on y-axis
  theme_bw() +
  geom_point() + # Plot the actual points
  #geom_point(aes(x=Mileage, y = y_bar), color="red", size=1) + 
  #geom_point(aes(x=Mileage, y = predicted), color="blue", size=1) + 
  geom_smooth(method="lm", se=FALSE, size=0.5)


```

---
# If the relationship were weak

```{r, warning=FALSE, message=FALSE, error=FALSE}
ggplot(weak, aes(x=x, y=y)) +  # Set up canvas with outcome variable on y-axis
  theme_bw() +
  geom_point() + # Plot the actual points
  #geom_point(aes(x=Mileage, y = y_bar), color="red", size=1) + 
  #geom_point(aes(x=Mileage, y = predicted), color="blue", size=1) + 
  geom_smooth(method="lm", se=FALSE, size=0.5) +  
  #geom_hline(yintercept=0, color="red") +
  geom_segment(aes(x=-2 ,xend=2,y=weak$y_bar,yend=weak$y_bar), color="red") +
  geom_segment(aes(x=x, y=y, xend = x, yend = predicted), color="blue", linetype="solid", size=0.5) +
  geom_segment(aes(x=x, y=predicted, xend = x, yend = weak$y_bar), color="red", linetype="solid", size=0.5) +
  annotate("text",x=-2.25, y=1.7, hjust=1, vjust=0.5,
            label=TeX("$y_i$"), size=8) +
  annotate("text",x=-2.25, y=-0.4, hjust=1, vjust=0.25,
            label=TeX("$\\hat{y}_i$"), color="blue", size=8) + 
  annotate("text",x=-2.25, y=0, hjust=1, vjust=0.5,
            label=TeX("$\\bar{y}$"), color="red", size=8) + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank())

```

---
# If the relationship were strong

```{r, warning=FALSE, error=FALSE, message=FALSE}

strong <- as.data.frame(mvrnorm(1000, mu = c(0,0), Sigma = matrix(c(1,0.83,0.83,1), ncol = 2),
               empirical = TRUE))
fitstrong <- lm(V1 ~ V2, data = strong)
strong <- strong %>%
  mutate(predicted = predict(fitstrong),
         residuals = residuals(fitstrong),
         y_bar = mean(V1)
    ) %>%
  rename(x=V2, y=V1)


ggplot(strong, aes(x=x, y=y)) +  # Set up canvas with outcome variable on y-axis
  theme_bw() +
  geom_point() + # Plot the actual points
  #geom_point(aes(x=Mileage, y = y_bar), color="red", size=1) + 
  #geom_point(aes(x=Mileage, y = predicted), color="blue", size=1) + 
  geom_smooth(method="lm", se=FALSE, size=0.5)

```

---
# If the relationship were strong

```{r, warning=FALSE, error=FALSE, message=FALSE}

ggplot(strong, aes(x=x, y=y)) +  # Set up canvas with outcome variable on y-axis
  theme_bw() +
  geom_point() + # Plot the actual points
  #geom_point(aes(x=Mileage, y = y_bar), color="red", size=1) + 
  #geom_point(aes(x=Mileage, y = predicted), color="blue", size=1) + 
  geom_smooth(method="lm", se=FALSE, size=0.5) +  
  #geom_hline(yintercept=0, color="red") +
  geom_segment(aes(x=-2 ,xend=2,y=strong$y_bar,yend=strong$y_bar), color="red") +
  geom_segment(aes(x=x, y=y, xend = x, yend = predicted), color="blue", linetype="solid", size=0.5) +
  geom_segment(aes(x=x, y=predicted, xend = x, yend = strong$y_bar), color="red", linetype="solid", size=0.5) +
  annotate("text",x=-2.25, y=1.7, hjust=1, vjust=0.5,
            label=TeX("$y_i$"), size=8) +
  annotate("text",x=-2.25, y=-0.4, hjust=1, vjust=0.25,
            label=TeX("$\\hat{y}_i$"), color="blue", size=8) + 
  annotate("text",x=-2.25, y=0, hjust=1, vjust=0.5,
            label=TeX("$\\bar{y}$"), color="red", size=8) + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank())

```


---
class: middle, inverse, center

# A "significant" Model will have:

# Big SSM, Small SSE

---
class: middle, inverse, center

# ANOVA Table

```{r}
data("AccordPrice")
m1<-lm(Price~Mileage, data=AccordPrice)
```

---
#
Rows and Columns of the ANOVA Table
In R, each slope coefficient has its own row in the Table. The residuals have their own row.

### Degrees of Freedom

### Sum of Square

### Mean Squares ( $\frac{Sum Squares}{df}$ )

### F value (test statistic: $F= \frac{MSModel}{MSError}$)
  - $H_0: \beta_1=0$ and $H_A: \beta_1 \ne 0$

### p-value for the F test

### [Shiny App - practice your F test intuition](https://gallery.shinyapps.io/dist_calc/)

---
### Recreate for Accord Price

The df and Sum Sq are:

_ | DF | Sum Sq | Mean Sq | F value | Pr(>F)
------ | --- | --- |  --- | --- | 
Mileage | 1 | 687.66 | | |
Residuals | 28 | 266.49 | | |

---
# R-Squared

$$ R^2 = \frac{SSModel}{SSTotal} = \frac{\sum (\hat{y}-\bar{y})^2}{\sum (y-\bar{y})^2}$$

Use the values from the ANOVA table below between Price and Milege to calculate $r^2$.
```{r, echo=TRUE}
anova(m1)
```

> $R^2$, or the percent of variation in the response variable explained by the model, is ...

> XX% of the variation in used Honda Accord Prices can be explained by the car's mileage.

---
### Residual/Regression Standard Error

$$\hat{\sigma_{\epsilon}} = \sqrt{\frac{\sum(y - \bar{y}) ^2}{n-2}} = \sqrt{\frac{SSE}{n-2}}$$

Use the values from the ANOVA table below between Price and Milege to calculate $r^2$.
```{r, warning=FALSE, error=FALSE, message=FALSE}
anova(m1)
```

> The regression standard error, or the average amount of error in the regression model, is ...

> The regression model predicting used Honda Accords' prices as a function of the car's mileage is off by, on average, $XXX.

---
### These values aren't saved in R, but you can re-create them.
```{r, echo=TRUE}
#SSTotal = SSModel + SSE
SSTotal <- var( AccordPrice$Price ) * (nrow(AccordPrice)-1)
SSE<- sum(m1$resid^2 )
SSModel<- SSTotal - SSE
c(SSTotal, SSE, SSModel)
```

```{r, echo=TRUE}
# R2
our_r2<-SSModel/SSTotal
our_r2
```

```{r, echo=TRUE}
# RSE
n2_df<-nrow(AccordPrice)-2
our_rse<-sqrt(SSE/n2_df)
our_rse
```

---
# Cautions about $R^2$

- What is a "big" $R^2$?

- You likely don't care unless predicting
    - Remember why we model: prediction, assessing relationships, testing differences
        - Investment in education and GDP growth
        - Wage differences by age
